%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANNEXES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex A --- Inside Out Debate Format: First Design Iteration}
\addcontentsline{toc}{section}{Annex A --- Inside Out Debate Format: First Design Iteration}

The \textbf{Inside Out} format is an experimental debate format, designed using best practices from tabletop RPG game design. It seeks to be a more \textit{fun}, \textit{performance-growth oriented}, and \textit{real-life skill-building} alternative to traditional competitive debate formats. Rather than winning arguments, debaters explore values, make meaningful decisions, and earn recognition for the skills they demonstrate.

\subsubsection{Team Structure}

Debaters are organised in four teams of two debaters each and are assigned positions within the debate at random:

\begin{itemize}
    \item \textbf{Team S (Subject):} Represents the Subject of the Motion --- the person who must face a decision. Must weigh all values and choose.
    \item \textbf{Teams A, B, and C (Value Defenders):} Each argues that a particular Value held by the Subject should be prioritised in the decision, and consequently, what the Subject should decide.
\end{itemize}

\subsubsection{The Motion}

The Motion outlines:

\begin{itemize}
    \item What is the decision that has to be made.
    \item What is the background of the Subject and other relevant practical factors that can be taken into account within the decision.
    \item What are the three Values of the Subject, each assigned to one Value Defender team.
\end{itemize}

It is important to note that the Motion does not prescribe to the Value Defenders a particular position on what decision the Subject has to make. They have to construct it themselves.

\subsubsection{Speech Order}

The four teams hold alternating speeches in the following order:

\begin{center}
A1, B1, C1, S1, A2, B2, C2, S2
\end{center}

Within their speeches, the Value Defenders argue for the relevance of their value, construct a specific decision that best honours it, and engage with the arguments forwarded by the other Value Defenders.

Within their speeches, the Subject team must outline a strong reasoning that engages with and explains the interplay between all arguments of the Value Defenders, picking a particular decision. Within S1 they pick a preliminary decision and within S2 a final decision.

\subsubsection{Evaluation}

The debate is overseen by an Evaluator, who performs two tasks:

\paragraph{Commendations.} The Evaluator awards Commendations to debaters. Commendations represent acknowledgements of useful skills, knowledges, or attitudes. A list of awardable commendations is constructed by the Chief Evaluators of the Debate Tournament, along with the total number of Commendations that Evaluators can offer within that Tournament. Example commendations include Believable Passion, Witty Solution, Argument Ingenuity, and Resilience (see Section~9 of this Annex for the full list).

Over time, debaters who gather a particular commendation at least three times earn a Badge. Gathering it additional times upgrades the badge, an unlimited number of times, but each upgrade requires gathering three more Commendations than last time.

\paragraph{Oral Feedback.} The Evaluator delivers an Oral Feedback session, in which they explain to each team the commendations they received, the reasoning behind them, and give advice on how each debater can improve in the future based on their performance.

\subsubsection{Example Motion}

\begin{quote}
You are the parent of a 16-year-old teenager. Your family is middle class and lives in the suburbia of a metropolis. You have higher education. You live in a country in which Higher Education is free, but there are only a few universities that are deemed as good. The year is 2026. You evaluate that your kid is talented in music, is a good leader and has had very good STEM skills, but is also quite distracted, does not have many friends and has failed to learn a foreign language. Your kid expressed a moderate interest in studying Musicology at university after they finish high school, but is also considering not going to higher education. You have a good relationship with your child. They asked you to suggest what they should do. What do you recommend your kid they should do after they finish high school?

\medskip

Three Beliefs that you hold, relevant for this decision, are:
\begin{itemize}
    \item In life, financial stability and independence is essential for a happy life.
    \item It is important to be a well-rounded, cultured individual.
    \item The people that you surround yourself with end up defining who you are.
\end{itemize}
\end{quote}

\subsubsection{Comparison with Traditional Formats}

The table below contrasts Inside Out with World Schools debate along six dimensions.

\begin{center}
\begin{tabular}{|l|p{5cm}|p{5cm}|}
\hline
\textbf{} & \textbf{World Schools} & \textbf{Inside Out} \\
\hline
Teams & 2 teams $\times$ 3 members & \textbf{4 teams $\times$ 2 members} \\
\hline
Sides & Proposition vs.\ Opposition & \textbf{3 Value Defenders + 1 Subject} \\
\hline
Motion style & ``This House\ldots'' policy/value motions & \textbf{Personal dilemma with background \& values} \\
\hline
Goal & Win the argument --- defeat the other side & \textbf{Explore values --- help the Subject decide} \\
\hline
Evaluation & Scored on Content, Style \& Strategy (points) & \textbf{Commendations for skills \& attitudes (badges)} \\
\hline
Outcome & Win / Loss + speaker rankings & \textbf{No win/loss --- growth-oriented feedback} \\
\hline
\end{tabular}
\end{center}

\medskip

\noindent\textbf{Key shift:} Inside Out moves from adversarial persuasion to collaborative deliberation. There is no winner --- instead, debaters earn commendations that recognise individual skills, and the Subject synthesises all perspectives into a reasoned decision.

\subsubsection{Player Instructions}

\paragraph{Value Defenders.}
\begin{itemize}
    \item Formulate arguments for why your value is \textbf{important, useful, and relevant} to this dilemma.
    \item Propose a \textbf{specific decision} that best honours your value --- argue why it is right and leads to good outcomes, especially through the lens of your value.
    \item Engage with other Defenders' \textbf{values}: show yours is more important, more relevant to the dilemma, or safer to follow.
    \item Engage with other Defenders' \textbf{decisions}: attack, support (claiming it maximises your value), or endorse with addendums --- choose your positioning wisely.
    \item Your core mission: \textbf{define what honouring your value looks like} in practice, and persuade the Subject it is the right path.
\end{itemize}

\paragraph{The Subject.}
\begin{itemize}
    \item \textbf{Enter without an opinion.} You are completely torn --- you are here to listen and be persuaded, not to argue.
    \item Listen as if you \textit{are} the Subject: \textbf{let the speeches move you} as they would move this person.
    \item Also take a mental step back: \textbf{evaluate how persuasive each argument would be} for the Subject, and adjust your lean accordingly.
    \item Take notes on arguments and their interplay --- record what the Defenders say, \textbf{not your own personal thoughts}.
    \item \textbf{Collaborate quietly} with your partner: you are both embodying the same person, not competing.
    \item In your speeches, share which decision you are leaning toward and \textbf{why}, weighing all arguments to the extent they were persuasive. You must pick a specific decision, not merely name a value.
\end{itemize}

\paragraph{The Evaluator.}
\begin{itemize}
    \item Analyse the debate and each player's performance. Decide which \textbf{commendations} to award --- aim for one per debater (the same commendation can go to more than one person).
    \item Approach the round as an \textbf{educator} helping each player grow, not as a judge declaring a winner.
    \item Open with a \textbf{narrative of the debate} from your outside perspective: what each team did and how it all unfolded.
    \item Engage directly with each team: ask what they \textbf{felt they did well} and what they found difficult. Share your view and add your own observations.
    \item Ask each team what they want to \textbf{do better next time}, and give concrete advice on how to get there.
    \item Close by \textbf{awarding commendations} to each player.
\end{itemize}

\subsubsection{Rituals}

\paragraph{Rolling In.}
\begin{enumerate}
    \item \textbf{Team commitment chant} --- Each team member states their commitment aloud before the debate begins:

    Subject: \textit{``I am [character name]. I am open to being moved.''}

    Each Defender: \textit{``I believe in [value]. I will argue from that belief.''}
\end{enumerate}

\paragraph{Rolling Out.}
\begin{enumerate}
    \item \textbf{Subject's thanks} --- The Subject team thanks each Defender individually: \textit{``Thank you for speaking your truth.''}
    \item \textbf{Defenders' thanks} --- Each Defender team thanks the Subject: \textit{``Thank you for listening and carrying the decision.''}
    \item \textbf{De-roling} --- Everyone stands. Each person states their real name and shakes hands with every participant: \textit{``I am [name]. Thank you for playing with me.''}
    \item \textbf{Bleed check} --- The Evaluator asks quietly: \textit{``Did anything in that debate feel close to home for anyone?''} --- a brief, optional moment to name any emotional residue before moving into feedback. No one is required to answer.
\end{enumerate}

\subsubsection{Full Commendation List}

Commendations are organised across five categories. Each commendation has three award tiers: \textit{Earned} (threshold quality), \textit{Strong} (clear excellence), and \textit{Exceptional} (reserved for outstanding, rare performances). Some commendations carry role restrictions (Defenders only, or Subject only) as noted.

\medskip
\noindent\textbf{Style} \textit{(6 commendations --- delivery, presence, and rhetorical quality)}

\paragraph{Believable Passion.} You were truly invested in your point, showing genuine, believable conviction and care for the value you defended. Distinct from Honest Connection (self-disclosure) and Broad Range (technical delivery control).
\begin{description}
    \item[Earned] Tone conveyed genuine care at key moments; voice quality shifted naturally, eye contact held at key claims, word choices felt spontaneous.
    \item[Strong] Conviction was felt in the substance of the speech; the room or the Subject visibly reacted.
    \item[Exceptional] The delivery made the dilemma feel real and worth deeply caring about. Reserve for speeches where the affective dimension was genuinely transformative.
\end{description}

\paragraph{Entertaining Discourse.} The speech was genuinely enjoyable through wit, lightness, or a well-timed observation that made the room smile or lean in. Distinct from Poetic Appeal (literary craft).
\begin{description}
    \item[Earned] At least one moment was distinctly enjoyable: a clever comparison, a light touch, or a well-timed observation.
    \item[Strong] Liveliness was woven into the argumentation consistently; humor or vividness served to make points land more effectively.
    \item[Exceptional] Entertainment and argument were seamlessly fused throughout, reaching the level of a sharp public talk where wit and substance were inseparable.
\end{description}

\paragraph{Broad Range.} Your delivery showed impressive stylistic range, varying tone, rhythm, tempo, and volume to serve the speech. Distinct from Believable Passion (authenticity of feeling); rewards technical delivery control.
\begin{description}
    \item[Earned] At least two noticeably different delivery registers were used; the speech did not feel flat.
    \item[Strong] Variation was deliberate and purposeful: shifts in pace or tone coincided with structural moments in the argument.
    \item[Exceptional] The full palette was deployed with mastery; the debater controlled the room's attention through delivery alone.
\end{description}

\paragraph{Honest Connection.} Something genuine about yourself entered the speech --- a real disclosure rather than a rhetorical move --- that created contact between the debater and the room. Distinct from Believable Passion (conviction); rewards self-disclosure and personal presence.
\begin{description}
    \item[Earned] At least one moment of genuine self-disclosure (an admission, first-person reflection, or personal statement of feeling) that felt real rather than strategic.
    \item[Strong] The personal register appeared at key structural moments; eye contact moved around the room, direct second-person address, or deliberate pause.
    \item[Exceptional] The speech was uniquely theirs; you felt the debater was genuinely present in the room with you.
\end{description}

\paragraph{Poetic Appeal.} Language was used with aesthetic care, where rhythm, imagery, or metaphor elevated the speech as a piece of writing, independent of its entertainment value. Distinct from Entertaining Discourse (enjoyment response); rewards literary and rhetorical craft.
\begin{description}
    \item[Earned] At least one passage was noticeably well-crafted: a sentence that would stand alone as good writing, or a rhythmic construction that made a point feel inevitable.
    \item[Strong] Poetic quality sustained across multiple moments; rhetoric served the argument.
    \item[Exceptional] The speech had genuine literary quality; specific lines are still in your head afterward. Language itself became part of the argument.
\end{description}

\paragraph{True Embodiment.} You convincingly inhabited your role, arguing and deciding from inside it rather than about it from outside. For the Subject: thinking, feeling, and deciding as the character. For Defenders: arguing from genuine conviction in the value.
\begin{description}
    \item[Earned] The role's reality guided the speech throughout; stated leanings tracked the actual arguments made.
    \item[Strong] The embodiment elevated the entire round; you could see the speaker being moved, reconsidering, weighing.
    \item[Exceptional] Not applicable; omit Exceptional tier for this commendation.
\end{description}

\medskip
\noindent\textbf{Strategy} \textit{(7 commendations --- positioning, framing, and tactical intelligence)}

\paragraph{Specific Topicality.} Arguments were sharply specific to this dilemma, this Subject, and these values, with no generic claims. Distinct from Advantageous Framing (evaluative lens); rewards the content of the arguments themselves.
\begin{description}
    \item[Earned] Most substantive claims grounded in the specifics of the motion: the Subject's background, the values as stated, the actual decision at stake.
    \item[Strong] Topicality was strategic; the debater drew on details others missed.
    \item[Exceptional] Every substantive move calibrated to this specific situation; the arguments would fall apart if the Subject or dilemma were changed.
\end{description}

\paragraph{Advantageous Framing.} You defined the terms of discussion in a way that benefited your position before others could. Distinct from Specific Topicality (argument content) and Helpful Guiding (neutral Subject navigation).
\begin{description}
    \item[Earned] Explicitly established a coherent evaluation framework relevant to their value's claim.
    \item[Strong] The framing did real structural work; other debaters had to address it, and the Subject engaged with it directly.
    \item[Exceptional] The framing was original and visibly influenced the course of the debate. Award rarely.
\end{description}

\paragraph{Witty Solution.} You proposed a concrete, value-informed solution to the dilemma, specific enough that it could only have come from someone defending your value. Distinct from Advantageous Framing (criteria); rewards the solution itself.
\begin{description}
    \item[Earned] The proposed solution was specific and thought through, demonstrably flowing from the defended value.
    \item[Strong] Notably smart: navigated a genuine tension between values, or revealed an angle others missed.
    \item[Exceptional] Elegant and surprising, more persuasive than alternatives precisely because of how the value was applied.
\end{description}

\paragraph{Brave Pivot.} You changed position or conceded a point honestly, and it made your case stronger.
\begin{description}
    \item[Earned] Explicitly acknowledged an opponent's point had merit, or adjusted position in response to what was said.
    \item[Strong] The pivot was strategic: conceding less important ground to better defend more important ground.
    \item[Exceptional] Changed their mind in real time under pressure, arguing why the pivot extends rather than contradicts the previous position. Award with corresponding rarity.
\end{description}

\paragraph{Efficient Alliance.} \textit{(Defenders only)} You recognized when another Defender's arguments complemented yours and built on that alignment.
\begin{description}
    \item[Earned] Acknowledged at least one point of genuine agreement with another Defender that strengthens the case for a decision.
    \item[Strong] Explicitly constructed an alliance: ``both Team A and I agree that X, which shows the Subject that the weight of argument points toward Y.''
\end{description}

\paragraph{Careful Positioning.} \textit{(Defenders only)} Your stance toward other Defenders' decisions was deliberate and chosen rather than reactive, with positive effect.
\begin{description}
    \item[Earned] Response to other Defenders' decisions was evidently chosen rather than accidental; the stance was coherently maintained.
    \item[Strong] Positioning was in service of a coherent strategy; you could reconstruct the reasoning behind every stance taken, forming a consistent picture.
\end{description}

\paragraph{Helpful Guiding.} \textit{(Subject only)} You helped the room navigate the structure of the debate clearly, accurately, and fairly. Distinct from Advantageous Framing; the Subject's role is neutral navigation, not partisan framing.
\begin{description}
    \item[Earned] At least one moment of genuine meta-level navigation, e.g.\ ``the key question before us is now X, because Y has been established and Z has been challenged.''
    \item[Strong] The guiding was structurally important, identifying a genuine turning point or consolidating arguments fairly to all sides.
\end{description}

\medskip
\noindent\textbf{Content} \textit{(9 commendations --- argumentation, analysis, and engagement with the debate)}

\paragraph{Word Economy.} You communicated clearly and efficiently; every minute of speaking time counted. Rewards form of expression, not argument substance.
\begin{description}
    \item[Earned] The speech did not waste significant time; arguments were stated without excessive preamble, redundant restatement, or filler.
    \item[Strong] Concision actively served clarity: complex ideas were expressed with precision; understood fully on first hearing.
    \item[Exceptional] The debater said something hard to say in few words; specific lines had a density of meaning that rewarded attention.
\end{description}

\paragraph{Rebuttal Ingenuity.} Your rebuttal identified a non-obvious flaw in an opponent's argument: a hidden assumption, internal tension, or unconsidered consequence. Distinct from Argument Ingenuity (novel constructive arguments) and Comprehensive Breadth (coverage).
\begin{description}
    \item[Earned] Identified a non-obvious weakness rather than the first available attack.
    \item[Strong] Engaged with the strongest version of the argument and demonstrated why it fails even there.
    \item[Exceptional] Changed the argumentative landscape; the rebuttal was difficult or impossible to satisfactorily respond to.
\end{description}

\paragraph{Rebuttal Prioritization.} You made deliberate, strategic choices about where to concentrate your rebuttal effort, and those choices were correct. Distinct from Comprehensive Breadth (coverage); rewards deliberate focus over breadth.
\begin{description}
    \item[Earned] Concentrated force on the most important clashes, explicitly or implicitly flagging the allocation.
    \item[Strong] Prioritization revealed genuine strategic intelligence; the choice of what to ignore was as deliberate as what was attacked.
\end{description}

\paragraph{Rebuttal Accuracy.} You represented opponents' arguments faithfully before responding --- no mischaracterization, no straw-manning. No Exceptional tier; accuracy is a threshold quality. Award only when faithful representation was genuinely non-trivial.
\begin{description}
    \item[Earned] Paraphrased opponents' arguments correctly; engaged with the most charitable understanding.
    \item[Strong] Engaged with specific moments in specific speeches; accuracy made the rebuttal more effective, not just fairer.
\end{description}

\paragraph{Robust Logic.} Your arguments were well-constructed: valid, plausible, and without significant logical gaps. Distinct from Impressive Argumentative Depth; logic asks whether the reasoning chain holds structurally.
\begin{description}
    \item[Earned] Main arguments had clear claim-warrant-impact structure; a reasonable person could follow without supplying missing steps.
    \item[Strong] Logic held up under challenge: no internal contradictions, no steps requiring the audience to take things on faith.
    \item[Exceptional] Genuinely tight; would require a sophisticated objection to dislodge; obvious objections anticipated within the argument itself.
\end{description}

\paragraph{Impressive Argumentative Depth.} You went beyond the surface, interrogating implications, mechanisms, or philosophical roots. Distinct from Robust Logic (structural soundness) and Argument Ingenuity (novelty).
\begin{description}
    \item[Earned] At least one argument went meaningfully beyond its initial claim, explaining why the value's demand is well-founded or what it would look like for the Subject to live by it.
    \item[Strong] Depth sustained throughout; consistently pushed past surface-level argumentation.
    \item[Exceptional] After the speech, you found yourself reconsidering something you had assumed you understood, or the argument illuminated a dimension of the value no other speech had touched.
\end{description}

\paragraph{Comprehensive Breadth.} You engaged with the full scope of the debate, including inconvenient arguments. Distinct from Rebuttal Ingenuity (sharpness) and Rebuttal Prioritization (deliberate focus); rewards attentiveness across the full debate.
\begin{description}
    \item[Earned] Addressed the most important arguments from all other Defenders, not only the weakest or most convenient.
    \item[Strong] Demonstrated careful listening throughout; the speech reflected the actual state of the debate.
    \item[Exceptional] Managed to synthetically yet explicitly cover every significant aspect of the debate.
\end{description}

\paragraph{Argument Ingenuity.} You produced an argument that was unexpected, non-obvious, and substantively illuminating. Distinct from Rebuttal Ingenuity (reactive) and Impressive Argumentative Depth (thorough development); rewards novelty and surprise.
\begin{description}
    \item[Earned] At least one argument was notably fresh: not typically heard in this kind of debate, logically coherent, and able to withstand an obvious first objection.
    \item[Strong] Ingenious in the full sense: first surprised you, then convinced you, revealing something about the value or dilemma that prior thinking had missed.
    \item[Exceptional] The argument changed the debate; others had to respond to it and it became a fixed reference point. Very rare, award accordingly.
\end{description}

\paragraph{Accurate Synthesis.} \textit{(Subject only)} You gave a fair, accurate, and well-organized account of the full state of the debate. Distinct from Comprehensive Breadth (Defenders' engagement); rewards the Subject for faithfully describing and weighing what happened.
\begin{description}
    \item[Earned] Correctly identified the main arguments from each Defender and did not misrepresent or omit them, even where unfavorable.
    \item[Strong] Showed genuine comprehension of the debate's structure: an account of how arguments interacted, where they clashed, and what the real decision points were.
    \item[Exceptional] Identified the strongest and weakest versions of each Defender's case and grounded the final decision in a genuinely comprehensive account of the round.
\end{description}

\medskip
\noindent\textbf{Fairplay} \textit{(3 commendations --- observe team dynamics \& debrief, not only speeches)}

\paragraph{Teammate Support.} You actively helped your partner formulate something difficult in a way that visibly contributed to the team's performance. Look for: visible collaboration during prep time; a second speaker building clearly on something the first struggled with. Award when the help was substantive and its impact visible. Ask in the debrief if not fully observable.

\paragraph{Teammate Solidarity.} You showed genuine support for your partner during a hard moment, without distancing yourself. Look for: body language and verbal behavior toward a partner after a difficult moment; a second speaker who picked up where a struggling first left off without implicitly throwing them under the bus. The difficult moment must have been real and the solidarity genuine. If ambiguous, ask in the debrief.

\paragraph{Strategic Problem-Solving.} You demonstrated intelligent, adaptive problem-solving in the face of strategic challenges in the round. Look for: a visible strategy adjustment that worked; a debater who recognized a weakness in their own case and addressed it explicitly; coordinated strategic adaptation across both speeches. Both identifying and solving the problem must be visible and shared between the two team members.

\medskip
\noindent\textbf{Growth Mentality} \textit{(4 commendations --- use the debrief; always name the specific moment in feedback)}

\paragraph{Applied Feedback.} You visibly tried to apply feedback from a previous round; imperfect application still earns this. Evidence may come from direct comparison with a prior round or from the debater's own account in the debrief. What matters is the genuine attempt rather than the outcome.

\paragraph{Out of Comfort Zone.} You attempted something genuinely difficult for you and made a real go of it. Rewards the attempt, not the result. Comfort zone is individual; use the debrief to identify what was genuinely challenging. Ask directly if needed.

\paragraph{Tracking.} You showed clear awareness of your own growth across debates: where you started, what you have worked on, and where you are heading. Distinct from Applied Feedback (applying specific feedback in one round); rewards the broader self-awareness of one's development arc over time. Evidence may come from the debrief: the debater can articulate what they have improved, name skills still being worked on, and describe their trajectory with accuracy.

\paragraph{Resilience.} You encountered a genuinely difficult moment and came out intact, or stronger. The difficult moment must have been real. Award for: a strong attack responded to with a composed second speech; a delivery stumble overcome mid-speech; an unexpected move handled with adaptability. Name the specific moment in feedback; this commendation must be concrete to be useful.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex B --- Inside Out Debate Format: Second Design Iteration}
\addcontentsline{toc}{section}{Annex B --- Inside Out Debate Format: Second Design Iteration}

The second design iteration is based on observations and participant feedback gathered during the first playtest (Jean Monet debate club, Bucharest, March 2026). The changes below supersede the corresponding elements in the first iteration (Annex A).

\subsubsection{Speech Order Changes}

The interstitial S1 speech between the first and second rounds is removed. In its place, two new elements are introduced:

\paragraph{S0 --- Subject Opening Speech.} A new opening speech before any Defender speaks. The Subject team introduces the dilemma and argues why it is genuinely hard: all three values are relevant, the decision is real, and there is no obvious right answer. This speech gives the debate proper gravitas before Defenders begin.

\paragraph{The Huddle.} A 10-minute interval inserted after C1 (and before A2). The Subject team may ask direct questions to any Defender team, grant them the floor to answer, and revoke their right to speak at any point. The Huddle allows the Subject to control the conversation toward resolving argumentative tensions, forcing arguments that need to interact to interact.

\paragraph{Thinking Time.} A 2-minute silent thinking period is added before S2, giving the Subject team time to integrate all arguments before delivering their final decision.

The new speech order is:

\begin{center}
S0, A1, B1, C1, [The Huddle], A2, B2, C2, [2-min thinking time], S2
\end{center}

\textbf{Points of Information (POIs)} are introduced during all speeches: any debater may signal to the current speaker to ask a brief question.

\subsubsection{Motion Changes}

\begin{itemize}
    \item The Subject character description in the motion should include a longer list of the Subject's personal beliefs and past decisions, to flesh out the character more fully and enable value-specific argumentation.
    \item Value Defenders should address the Subject in the second person (``you, as the Subject, should\ldots'') rather than in the third person, reinforcing the personal and direct character of the deliberation.
\end{itemize}

\subsubsection{Evaluation Changes}

Commendations are awarded in private: after the debate, the Evaluator approaches each team separately rather than delivering commendations in a public session. The oral feedback narrative and growth-oriented debrief remain public.

\subsubsection{Speech Instruction Clarifications}

\begin{itemize}
    \item Subject instructions explicitly state that the Subject must pick a specific \textit{decision} (a concrete course of action), not merely name a value they find most important.
    \item More explicit guidance is provided on what the Subject team should prepare during prep time: for S0, they should prepare to explain the dilemma's genuine difficulty and why all values are relevant; for S2, they should prepare to synthesise all arguments and justify a final specific decision.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex C --- Debrief Form}
\addcontentsline{toc}{section}{Annex C --- Debrief Form}

The following questionnaire was administered to participants after the first playtest. Questions aim to gather data on entertainment value, format distinctiveness, participant intentions, and transformative effects.

\begin{enumerate}
    \item How fun was the debate? (scale 1--10)
    \item How fun was it compared to a debate in a traditional format? (5-choice scale)
    \item Was there anything surprisingly entertaining or engaging about it?
    \item Was there anything about the format that was quite difficult, to the point of frustration?
    \item Was there anything about the rules of the format that was confusing?
    \item Would you play this format some more? (Yes / No)
    \item If you would play it some more, in what circumstance would you?
    \item If there was to exist an alternative debate circuit centred around this format, would you attend those competitions? Even at the expense of competitions in the WS or BP format?
    \item Do you believe you played differently than you would have in a debate in WS/BP with a rather similar motion? If so, how?
    \item Do you feel different than how you usually feel after a debate? If so, how?
    \item Did you perceive yourself and others speaking similarly or differently than how you do in a regular debate? If so, how?
    \item Were you to redo this debate, what would you do differently?
    \item Were you to play more in this format, but on different motions, what would you do differently?
    \item What do you believe you learned from this debate?
    \item What do you believe this debate format does better, as a learning/empowerment tool, than WS or BP? Do you believe it does something worse?
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex D --- Playtest Motion}
\addcontentsline{toc}{section}{Annex D --- Playtest Motion}

The following motion was used in the first playtest at the Jean Monet debate club, Bucharest, March 2026.

\paragraph{First version.} Used in the first playtest at the Jean Monet debate club, Bucharest, March 2026.

\begin{quote}
You are Alex. You are a lawyer. You are close to your older brother Matei. He is a very caring person to everyone, always eager to help even strangers. He always helped you when you were in need. However he was always quite dishonest with you, and you caught him lying even about important things. Matei is also an alcoholic and you believe he is quite bigoted.

\medskip

Matei was involved in a hit and run recently. He hit a cyclist that ended up hospitalised. The cyclist remembered his license plate only partially and he is now being investigated by the police. From your knowledge they only have circumstantial evidence, but he still risks going to jail. He asked you to lie that he was with you at home at the time of the accident, in order to provide him an alibi. Your assessment as a lawyer is that this will almost certainly exonerate him. However, a very small risk remains that additional material evidence could be found, at which point you might become an accessory to the crime.

\medskip

\textbf{What should you do?}

\medskip

Three Values that you hold, relevant for this decision, are:
\begin{itemize}
    \item You believe that helping people in need is very important, and a moral duty.
    \item You believe in justice: that people should receive what they deserve.
    \item You believe that family should stick together and have each other's back.
\end{itemize}
\end{quote}

\paragraph{Second version.} Used in the second playtest, April 2026.

\begin{quote}
Alex is a criminal defence lawyer with seven years of experience. The eldest of three siblings from a small town, Alex was the first in the family to attend law school. After graduating, Alex spent two years at a legal aid clinic representing clients who could not afford private counsel, before joining a private firm. Alex's first major case there was defending a driver charged with causing serious injury in a road accident --- Alex secured his acquittal.

\medskip

Matei is Alex's older brother and the two have remained close throughout their lives. Matei is known among everyone who knows him for being eager to help others, often going out of his way even for strangers. He helped Alex through several difficult periods. He has also been dishonest with Alex repeatedly over the years, including about important matters. Matei struggles with alcohol and holds a number of views that Alex finds bigoted.

\medskip

Matei was recently involved in a hit-and-run: he struck a cyclist who was hospitalised. The cyclist recalled his license plate only partially, and Matei is now under police investigation. As far as Alex knows, the evidence against him is circumstantial. Matei has asked Alex to lie and say he was at home with Alex at the time of the accident. Alex's legal assessment is that this would almost certainly exonerate him, though a small risk remains that additional material evidence could surface --- at which point Alex would become an accessory to the crime. \textbf{What should you do?}

\medskip

Values:
\begin{enumerate}
    \item You believe that helping people in need is very important, and a moral duty.
    \item You believe in justice, that people should receive what they deserve.
    \item You believe that family should stick together and have each other's back.
\end{enumerate}
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex E --- Informed Consent Form}
\addcontentsline{toc}{section}{Annex E --- Informed Consent Form}

\bigskip

I, \underline{\hspace{7cm}} of \underline{\hspace{1.5cm}} years of age, voluntarily accept to participate in a playtest followed by an interview session, both moderated by Andrei Dan Olaru, as part of his \textit{``Applying Transformative Game Design Theory to Competitive Debate''} study. This study seeks to explore the creation of an alternative debate format, informed by modern transformative game design theory. I understand that:

\begin{enumerate}
    \item This session implies video recording of the entire debate playtest and the subsequent collective interview, plus some additional questions using an online questionnaire.
    \item Pseudonyms will be used when analysing the data for confidentiality, and the researchers commit to not revealing the identities of the participants. Individual raw data may be discussed in class or with the supervisor without the participant's name to trace back to. The data obtained will be analysed by grouping and some textual examples will be used anonymously.
    \item The recording will be safely stored in the university's OneDrive to maintain the saved raw interviews protected. The video will be erased one month after the dissertation's results. The transcripts will be erased up to one year after the dissertation's results.
    \item None of the questions will go against my integrity, no harm will come to me during this session, no answer is right or wrong and I may choose not to answer any question(s) without consequences of any kind.
    \item I am free to walk out of the playtest and interview at any time if I deem it necessary without prejudice or any consequences.
\end{enumerate}

For any additional information or difficulty, the participant may contact the researcher Andrei Dan Olaru at \texttt{andrei-dan.olaru.0744@student.uu.se}.

\bigskip\bigskip

\noindent
\begin{tabular}{p{6cm} p{1cm} p{6cm}}
\underline{\hspace{6cm}} & & \underline{\hspace{6cm}} \\
Participant's signature & & Researcher's signature \\
\end{tabular}

\bigskip\bigskip

I, \underline{\hspace{7cm}} have been informed of the conditions and freely accept to participate in the research by Andrei Dan Olaru.

\bigskip

\noindent
\begin{tabular}{p{6cm} p{5cm}}
\underline{\hspace{6cm}} & \underline{\hspace{3cm}} of March 2026 \\
Participant's signature & Date \\
\end{tabular}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex F --- Second Playtest Report}
\addcontentsline{toc}{section}{Annex F --- Second Playtest Report}

The following is a first-person account of the second playtest session, held in April 2026. The session gathered seven veteran debaters. Design changes arising from this playtest are documented separately in Annex G.

\subsubsection{Setting}

I held the second playtest together with seven veteran debaters, close friends. We held it at the apartment that part of us commonly rent and share with other nerds for the single purpose of playing TTRPGs there. Because of that, the main room is dominated by a hilariously large, roughly round table. We held the debate seated at it, which one player noted created a council-feel to the whole session --- as opposed to a regular debate in which the seating mimics parliamentary discourse and thus feels profoundly more impersonal. I was not the Evaluator this time; I played as the second speaker of the A Inner Voice team.

\subsubsection{The Brief}

The brief took about 40 minutes --- similar to how much it took with the high-schoolers. The fact that it required that amount of time even with people accustomed to parsing lengthy rulebooks speaks to how much the design differs from traditional formats. The division of the ``judge'' role into two separate individuals (the Subject and the Evaluator) was noted as interesting by several players. The role of the Subject was by far the hardest for players to understand, especially the purpose of the opening Subject speech. I had, at the point of writing the rules, found it genuinely difficult to pin down what the Subject should do in that first speech. I wrote vague guidelines --- and this showed --- but in the end I believe it was for the better: I let the player decide what would be most interesting and useful. In that I was fortunate: the role was drawn by a player with substantial game-design experience, a professional L\&D background, many years as a trainer, and deep experience as a roleplayer.

\subsubsection{Rituals}

The roll-in rituals were experienced as cringe, and were surely too lengthy. Players felt they should be replaced by a short introduction from the Evaluator followed by a brief shared gesture or chant. This aligns with how regular debates open: the judge makes an introduction, then everyone bangs the table vigorously and the debate begins. One player noted that the rituals do recognise the effort put in by the judge --- in this case shared between the Subject and the Evaluator.

The calling rituals were received positively.

The roll-out rituals were also too lengthy and became cringe, but one player noted they liked the way in which people express gratitude to everyone. The consensus was that this could be replaced by a moment of handshakes in which each person says a short phrase based on their role. Players also noted that calling a defender by name during the calling ritual felt unnecessary.

\subsubsection{S0 --- The Opening Subject Speech}

The player who drew the Subject role delivered an interesting, roleplay-heavy speech --- they even played a soundtrack. They fleshed out the motion with many evocative details that further emotionally contextualised the scenario, then tried to explain how they care about all three values and why each is relevant to the decision. This in part exposed some nuanced personal biases that bled from the player into the character. Several players --- including that player --- noted that this institutionalises the already existing practice of debaters ``playing for the judge'': attempting to exploit the judge's known biases. I found this observation striking. The speech, although I was sceptical of it initially, does hold a great deal of value: fleshing out the scenario, giving a floor for surfacing biases. What it needs is clearer language for what players are supposed to do.

\subsubsection{First Round of Inner Voices}

All players immediately adopted a roleplay stance. They spoke directly to the Subject, addressing them by name and in the second person, looking at them, and choosing speech patterns that mirrored those the Subject had displayed. They also took the liberty to add small details to the motion --- to worldbuild.

I find this very promising but also somewhat concerning. Giving players unofficial worldbuilding powers may lead inexperienced players to twist the debate in an unhelpful direction. However, adding colour to the broad facts of the scenario does lend to a more realistic simulation of how people actually argue, raises emotional engagement, and heightens immersion.

It was interesting that even very veteran debaters struggled, especially in this first round, to both appeal to emotion and to argue specifically for their assigned value. Most focused on constructing interesting and nuanced Solutions and arguing why these were practical and useful --- utilitarian and cerebral overall --- except for the last team, who appealed explicitly to emotion, which set the tone interestingly for the second half.

Players also began referring to each other as ``the other inner voices'' --- an organic renaming that feels more natural than ``Value Defenders''. I plan to adopt it.

\subsubsection{The Huddle}

Everyone found the Huddle very, very fun. The Subject player later said it was a genuine eureka moment when they realised they could ask particular Inner Voices to have very specific conversations with each other. It is worth noting that this player has meeting facilitation experience. During the Huddle, the Subject first asked each team to clarify things they were still confused about, then asked teams to expand on arguments they found particularly interesting, then asked pairs of Inner Voices to directly clash on issues they felt needed resolving.

At a certain point the Subject ran out of ideas for what to ask. I believe that with two players on the Subject team this would happen less frequently. Still, ten minutes felt too long --- I would cut it to eight.

\subsubsection{Second Round of Inner Voices}

The Huddle exacerbated something that had already started in the first round: a slow drift toward consensus on what decision should be taken. The second round was much less about the practicalities of which decision is best, and much more about arguing which is the dominant lens through which the decision should be viewed.

My intuition is that something similar might happen with other groups, not only with veteran debaters.

During my own speech I also pivoted toward consensus on the right Solution. I did this because I found it interesting and novel --- something regular debate never allows --- and I wanted to see how it felt, to demonstrate that the rules permit it, and to explore whether I could deliberately fish for the ``Brave Pivot'' commendation. I found that I felt little need to rebut the points made by others; arguing for my lens and shifting focus felt like a more natural and persuasive strategy than attacking others, and I allowed myself to be openly persuaded by fellow Inner Voices on some points.

A fellow player later noted that in this format to outright attack what others say ``feels wrong'' --- disrespectful and unnecessary --- and that the format requires tact in how one goes about diminishing the relevance of another's points.

\subsubsection{Second Subject Speech}

The Subject player structured their speech quite like a regular debate adjudication: they explained what they understood each Inner Voice team to have argued and how convincing each case was, then expressed the solution they had reached and what had factored most in their decision. They later said that during the first speech they were in more of a roleplay mindframe, but that as the debate progressed they transitioned more and more to feeling and thinking like a debate judge.

My intention had been for the second speech to remain somewhat more roleplay-ish --- the Subject speaking about how something \textit{felt} persuasive, being a bit less analytical. I am not sure whether adjusting the instructions is needed, or whether this was simply the style of this particular player.

\subsubsection{The Evaluator Speech}

The Evaluator gave insightful feedback on how each of us could have been more persuasive to a regular person. She then noted it was very difficult to find feedback to give, because the format does not specify clearly toward what educational objectives the feedback should be directed.

This raised a broader question I should think about more: traditional debate exists within a complex, organic constellation of clubs and competitions. Most people tell themselves they do debate to learn --- critical thinking, public speaking --- and trainers and judges tell themselves they are there to teach useful skills. In practice, my sense is that the circuit has a life of its own, and that people invest effort on either side mostly for the social reputation it yields.

How should this format be played? Who should facilitate it? Who is it for? What should it help participants achieve?

Another interesting suggestion was that a structure allowing the Evaluator to be more actively involved --- to facilitate and give real-time feedback, to encourage players --- would be useful. I tend to agree. Renaming the Evaluator ``the Facilitator'' might help signal this.

The Evaluator also said that awarding commendations was difficult, and that in a larger circuit it would become easier: she would not feel pressure to give everyone a commendation and would reserve them for truly outstanding moments. She added that once the commendations were memorised the system would feel much more organic than the numeric speech-point quantification that most traditional formats require of judges.

\subsubsection{On Bleed}

I decided to skip the final bleed-releasing ritual, expecting it would feel awkward. It was interesting, then, that even without it, one of the players spontaneously shared during a brief intermission that the debate had hit uncannily close to home: on the way to the session, a family member had asked them to testify for them in court, which they had found morally wrong.

\subsubsection{On Commendations}

There was substantial discussion about the commendation system. The more ``gamer'' of the players expressed that it scratches a ``make a build'' itch, but felt that commendations should carry some bonuses as well. They also raised the concern that players might start farming them --- choosing to play in unusual ways simply to collect a specific commendation and ``complete'' a build.

\subsubsection{Participant Feedback}

Most players expressed that the game was quite fun and that they believed it would be enjoyable even outside a casual friends setting.

One noteworthy pattern in the feedback: women and players who had been less successful as traditional debaters said they strongly preferred this format. Men who had been highly successful debaters expressed finding it less exhilarating. One participant put it directly: ``There are few things more ego-boosting than the feeling of total mind-domination that you get when you win a debate, and this cannot give you that.''

One point everyone agreed on was that the format is much less anxiety-inducing than regular debate, while still providing a structure for meaningful conversation --- unlike open real-life discussions, where people tend to fight over talking space.

One player suggested that the format might benefit from transitioning more wholeheartedly toward improv theatre --- becoming a roleplaying game without any performance-evaluation component. Another countered that there is value in the sense that you are always progressing, at least a little, with every debate.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex G --- Inside Out Debate Format: Third Design Iteration}
\addcontentsline{toc}{section}{Annex G --- Inside Out Debate Format: Third Design Iteration}

The third design iteration is based on observations and participant feedback gathered during the second playtest (April 2026). The changes below supersede the corresponding elements in the second iteration (Annex B).

\subsubsection{Role Renaming}

\textbf{Value Defenders are renamed Inner Voices.} This name emerged organically during the second playtest, when players began referring to each other as ``the other inner voices''. It better reflects the roleplay-oriented nature of the format and the relationship between the Value Defender teams and the Subject.

\subsubsection{S0 --- Opening Subject Speech Instructions}

The instructions for the opening Subject speech are revised to foreground the roleplay dimension:

\begin{quote}
Hold a monologue as if you are the character, alone in a place of comfort, speaking to yourself about the dilemma. Express why this decision is truly difficult and how much getting it right matters to you. Explain why each value is relevant to this decision, and how you cannot simply disregard any of them. Do not lean toward any decision yet.
\end{quote}

\subsubsection{The Huddle}

The Huddle is shortened from 10 minutes to \textbf{8 minutes}.

\subsubsection{S2 --- Final Subject Speech Instructions}

The instructions for the final Subject speech are revised to include emotional closure:

\begin{quote}
Synthesise all arguments and deliver a decision over a specific solution --- a concrete course of action from those presented by the Inner Voices. Explain how everything in the debate contributed to this decision; leave nothing important out. Say why you feel this is the right decision, and how you now feel about making it.
\end{quote}

\subsubsection{Seating}

A seating instruction is added: try to arrange chairs so that the Inner Voices and the Subject face each other. Inner Voices speak from where they sit, standing or seated. The Subject does the same. The Evaluator sits somewhere outside the main circle --- present but out of sight, out of mind.

\subsubsection{Specific Times}

The following timing guidelines are provided:

\begin{itemize}
    \item \textbf{Speaking times:} 5 minutes per speech.
    \item \textbf{Prep time:} 30 minutes for beginners or high-school debaters; 15 minutes for experienced adults.
    \item \textbf{Evaluator oral critique:} 10 minutes.
    \item \textbf{Evaluator-led debrief:} 15 minutes.
\end{itemize}

\subsubsection{Rituals}

The following rituals replace those described in the first design iteration (Annex A).

\paragraph{Rolling In.}
\begin{enumerate}
    \item \textbf{Evaluator opens} --- The Evaluator says: \textit{``You are here to decide together.''}
    \item \textbf{All respond} --- Everyone answers: \textit{``Yes, we are.''} followed by shared clapping.
\end{enumerate}

\paragraph{Calling an Inner Voice.}
\begin{enumerate}
    \item The Evaluator calls: \textit{``Voice, speak for [Value].''}
    \item The speaker \textbf{must begin their speech} with the words \textit{``I will\ldots''}
\end{enumerate}

\paragraph{Rolling Out.}
\begin{enumerate}
    \item \textbf{Voices' thanks} --- Each Inner Voice says: \textit{``Thank you for listening to me.''}
    \item \textbf{Subject's thanks} --- The Subject says to the Inner Voices: \textit{``Thank you for speaking your truth.''} The Subject then says to the Evaluator: \textit{``Thank you for listening to me.''}
    \item \textbf{Bleed check} --- Everyone states their real name and how they feel right now. The Evaluator then leads a brief debrief.
\end{enumerate}

\subsubsection{Commendation System}

The commendation system is updated with a tier structure governing how commendations are distributed:

\begin{itemize}
    \item Every debater should receive at least one \textbf{T1 (Earned)} commendation per speech. At higher levels of play, T1 commendations carry less competitive weight but remain a baseline acknowledgement of participation.
    \item The Evaluator has a \textbf{limited pool of T2 (Strong) and T3 (Exceptional)} commendations to offer per debate. These are awarded in private.
    \item Some commendations may be \textbf{exclusive to specific Evaluators}, reflecting their expertise or perspective.
    \item Some rounds or competitions may designate a \textbf{featured commendation} --- a particular commendation given special significance in that context.
\end{itemize}

\subsubsection{Evaluator Instructions}

The Evaluator role is reframed toward active facilitation:

\begin{itemize}
    \item You are the \textbf{facilitator} of the entire event. Your primary job is for everyone to feel safe and to play to grow.
    \item Feel free to \textbf{intervene} when it feels natural. Encourage players when you think it is necessary. Gently redirect when someone may be making others uncomfortable. Invite players out of their comfort zone when you believe it is in their best interest. Be human; establish a relationship of trust and growth with the others.
    \item After the debate ends, \textbf{first facilitate a brief debrief} stemming from the bleed check moment, then transition to the oral critique.
    \item Prepare \textbf{potential debrief questions} whilst others are preparing for the debate. Debrief questions should \textit{not} focus (as the oral critique does) on skill growth, but rather on \textbf{meaningful emotional growth and acceptance of others' realities}. Explore how debaters would act in a similar scenario and what their own values are regarding the values explored in the debate. When in doubt, use these two anchor questions: \textit{``How important are these values for you?''} and \textit{``Were you in a similar scenario, what other values would influence you?''}
    \item Open the oral critique with a \textbf{narrative of the debate} from your outside perspective: what each team did and how it all unfolded.
    \item Engage directly with each team: ask what they \textbf{felt they did well} and what they found difficult. Share your view and add your own observations.
    \item Ask each team what they want to \textbf{do better next time}, and give concrete advice on how to get there.
    \item Award at least one \textbf{T1 (Earned)} commendation per speech to every debater. Use your limited pool of \textbf{T2 (Strong) and T3 (Exceptional)} commendations sparingly --- these are awarded in private.
\end{itemize}
