\section{BACKGROUND AND LITERATURE REVIEW}

\subsection{Topic Overview}

\subsection{Debate as Transformative Game}

Academic debate formats are games. Following \textcite{salen2004rules}, a game is a system in which players engage in artificial conflict, defined by rules, resulting in a quantifiable outcome. Academic debate satisfies all criteria: rules, participants, winners, losers, tournaments, and scoring systems. Alfred Snider's 1982 doctoral dissertation at the University of Kansas, along with his subsequent publication \parencite{snider1984gaming}, established the scholarly foundation for viewing debate as a game while documenting its educational benefits.

But what makes a game \textit{transformative}? \textcite{bowman2025transformative} conceptualize transformative role-playing games as those intentionally designed to encourage personal and social change arising from play. For complete transformation to occur, this change must expand beyond the bounds of the original experience and integrate into the participant's daily frames of reality and identity. \textcite{daniau2016transformative} conceptualizes this transfer as one ``from play skills to human skills,'' showing through action-research how participants in transformative role-playing games engage across four levels of reality (character, player, person, and human being) and, through structured debriefing, convert in-game abilities into lasting personal competencies. By this standard, a game is truly transformative not merely when participants report enjoyment or in-game learning, but when the experience produces lasting changes in cognition, behavior, or self-concept. The question for academic debate, then, is whether it meets this threshold.

Forensics pedagogy documents debate's transformative potential across several domains that satisfy this criterion. The most extensively studied is critical thinking: debaters develop the ability to recognize stated and unstated assumptions, evaluate causal inferences, and assess the validity of arguments \parencite{colbert1995enhancing, hill1993value}. Beyond cognitive gains, debate serves as a vehicle for empowerment, particularly for previously ``silenced student populations'' who develop their own voice through the activity \parencite{snider2003gamemaster}, with Urban Debate Leagues leveraging this potential in academically challenged environments. The activity cultivates what \textcite{hicks1998public} calls a ``democratic ethos'': the capacity to justify convictions in public, respect minority views, and evaluate opposing arguments critically. Oral communication proficiency develops through audience adaptation, where students learn to analyze listeners' beliefs and norms to frame compelling narratives \parencite{cirlin1997sociological}, while parliamentary formats foster extemporaneous thinking and speaking \parencite{whitmore1998british}. Debate also functions as a ``laboratory'' for intensive research and information management \parencite{brodabahm2002perspectives}, with the intensity of high-level preparation frequently compared to master's thesis work \parencite{combs2002applying}. Finally, debate provides a socialization mechanism, integrating students into a sub-culture with specialized language that enhances their sense of belonging to an intellectual community \parencite{cirlin1997sociological, kargacin1998state}. Crucially, these effects are not confined to the debate round itself. They transfer into participants' academic performance, professional communication, and civic engagement, satisfying the criterion of transformation beyond the game.

Empirical research substantiates these claims, with quantitative evidence accumulating across critical thinking, academic achievement, and college readiness over several decades. \textcite{allen1999meta} conducted a meta-analysis of studies examining forensics and communication education, finding a significant positive relationship between such training and gains in critical thinking (longitudinal designs $r = .18$, cross-sectional designs $r = .20$), with forensics participation demonstrating the largest effect among the communication activities studied. At the individual study level, \textcite{colbert1987effects} conducted the most comprehensive investigation, tracking 275 CEDA and NDT debaters and non-debaters across nine universities over a full academic year. Using analysis of covariance controlling for pretest scores, he found that debaters' Watson-Glaser Critical Thinking Appraisal scores rose from 61.18 to 64.53 while non-debaters' scores actually declined from 52.67 to 49.14, with NDT debaters significantly outgaining non-debaters ($F = 31.77$, $p < .001$). While \textcite{hill1993value} cautioned that the broader body of debate-critical thinking research remained inconclusive due to methodological limitations and inconsistent results across sub-samples, \textcite{colbert1995enhancing} responded that the preponderance of evidence establishes ``presumptive proof'' favoring a positive debate-critical thinking relationship, particularly when properly designed studies using robust statistical controls are considered.

Beyond critical thinking, debate participation correlates with substantial academic outcomes. \textcite{mezuk2009urban} analyzed data from the Chicago Debate League spanning 1997 to 2006 and found that participants were 70\% more likely to graduate and three times less likely to drop out than comparable non-participants, even after controlling for eighth-grade test scores and grade point average. Debaters were also more likely to meet ACT benchmarks for college readiness in English and reading, though not in science or mathematics. A subsequent expanded study of the same program by \textcite{mezuk2011impact}, analyzing 9,145 students of whom 2,449 participated in at least one CDL tournament, found that debaters were 2.5 times less likely to drop out, graduated at a rate of 90\% compared to 75\% for non-debaters, and achieved a mean twelfth-grade GPA of 3.23 compared to 2.83 for their peers. Most recently, \textcite{schueler2025debate} examined policy debate in Boston Public Schools serving large concentrations of economically disadvantaged students and found positive impacts on English Language Arts scores of 0.13 standard deviations, equivalent to 68\% of a full year of average ninth-grade learning. Critically, these gains were concentrated on analytical rather than rote subskills and were largest among the lowest-performing students at baseline, with additional evidence of positive effects on high school graduation and postsecondary enrollment.

Forensics scholars have proposed several theoretical frameworks to explain how debate produces these transformative effects, each grounded in established learning theory. At the cognitive level, \textcite{colbert1995enhancing} argues that critical thinking develops through what he calls ``intellectual disequilibrium,'' a concept rooted in Piaget's constructivism, where confrontation with opposing positions forces learners to restructure their mental models rather than simply absorb new information. Debate's adversarial structure ensures this confrontation is unavoidable: a debater cannot ignore the opposing case but must engage with it directly, producing the cognitive conflict that drives accommodation. At the communicative level, \textcite{cirlin1997sociological} draws on Richard Weaver's rhetorical theory to show that debate develops persuasive skill through audience adaptation: debaters must continuously analyze their listeners' beliefs and norms to construct compelling narratives, a practice that transfers to communication contexts outside the activity. At the social level, Cirlin provides a sociological account of how participation integrates students into a discourse community with shared analytical vocabulary and norms of evidence evaluation. This linguistic socialization, which Cirlin describes through Weaver's concept of ``sermonic'' language, reshapes how participants perceive and engage with arguments in ordinary life. At the civic level, \textcite{hicks1998public} argues, drawing on Rawls's ideal of public reason, that debate cultivates a ``democratic ethos'': the disposition to justify convictions through publicly accessible reasoning, to respect minority views, and to evaluate opposing arguments on their merits. Drawing these threads together, \textcite{bellon2000research} synthesized evidence from competitive forensics, communication pedagogy, and educational psychology, including research on active learning and the social construction of knowledge, to argue that debate is a uniquely effective pedagogical tool precisely because it engages all of these levels simultaneously. However, these forensics-specific frameworks have not been connected to broader theories of experiential and transformative learning. In the Theoretical Framework section of this paper, I apply \textcite{vygotsky1978mind}'s zone of proximal development to the scaffolded learning that occurs through debate coaching and graduated competition, map \textcite{kolb1984experiential}'s experiential learning cycle onto the iterative process of debating, receiving feedback, and refining strategy, and draw on research into structured debriefing in simulation and role-playing \parencite{crookall2014engaging, daniau2016transformative} to illuminate how debate's built-in oral critique enables the reflective processing essential to lasting transformation.

Although forensics pedagogy has successfully documented debate's transformative effects and proposed explanatory mechanisms, I have yet to find any paper that analyzes major debate formats from a game design perspective or presents academic debating as a role-playing game, though \textcite{bowman2014educational} acknowledges related pedagogical approaches such as process drama, classroom debates and simulation as role-playing games. Transformative role-playing game theory provides sophisticated conceptual tools for understanding how structured play produces personal change, yet these have not been applied to competitive debate. Addressing this gap in literature, applying game design lenses and expertise to the analysis and improvement of debate formats, is the main contribution to knowledge that I seek to start making through this paper.

\subsection{Debate as Role-Playing Game}

In the World Schools format, the one I will most focus on in this paper, in every debate there are two teams which receive a motion (e.g., ``This House Would legalize the possession of drugs''). Teams are randomly assigned to opposite sides: Proposition/Government or Opposition. They deliver alternating speeches as if they were individuals supporting or opposing the motion. Judges listen as someone with no opinion on the motion, expert in debate rules, and averagely informed, deciding who most convinced the character they role-play. The debate ends with the judge ruling who won, explaining their decision and then offering feedback for improvement to the teams.

Understanding debate's relationship to role-playing games requires first establishing what role-playing games are and where debate fits among related practices. As \textcite{zagal2024definitions} observe, defining role-playing games is notoriously contentious, with scholars, designers, and practitioners disagreeing across forms, styles, and communities. However, certain characteristics commonly recur: role-playing games are play activities revolving around the rule-structured creation and enactment of characters in a fictional world, where players define and pursue their own goals with considerable choice in what actions they can attempt \parencite{zagal2024definitions}. \textcite{bowman2025transformative} offer a functional definition: analog role-playing games are ``co-creative experiences in which participants immerse into fictional characters and realities for a bounded period of time and improvise through spontaneous, emergent playfulness'' (p. 16).

Transformative game design literature explicitly positions debate among ``cousin forms'' of analog role-playing games \parencite{bowman2014educational,bowman2025transformative}, alongside simulations, educational role-playing, improv theatre, and psychodrama. The distinction rests not on the act of role-playing itself, which is much the same regardless of community, but on practices, norms, and innovations developing within each community. For the analytical purpose of usefully employing the body of theory developed to explain how and how better can role-playing games harness positive transformative energies, this is already enough to justify the following section. As will be later revealed, most of the theory developed within the Nordic larp tradition explains processes common to all role-playing activities.  Going even further however, competitive debate occupies an unusual position among these cousin forms, one that warrants closer examination. While debate is indeed used in educational contexts, it differs fundamentally from close cousins such as work simulations or classroom exercises in ways that bring it closer to analog role-playing games proper. Three theoretical dimensions illuminate this proximity: the depth of character immersion, the strength of alibi, and the leisure-community dimension.

First, regarding immersion: As opposed to work simulations or boardgames, debate requires players to genuinely inhabit an epistemic stance, constructing and defending arguments as if they believed them. This process of active perspective-taking requires what \textcite{bowman2010functions} describes as adopting a theory of mind, thinking ``as if'' one were someone else in a unique set of circumstances (p.~57). The debater must anticipate objections, identify weaknesses, and respond to challenges as someone who holds the assigned position, a cognitive demand that approaches the immersion into character central to role-playing games. As \textcite{seo2022good} observes, ``the moment one received a motion, one adopted the mindset of a person who was completely convinced of that point of view. One clung to this feeling of absolute conviction to make arguments, sink objections, and display passion'' (p.~222). Rules also encourage speaking passionately: \textcite{snider1999code} instructs debaters to speak ``with energy, enthusiasm, commitment, and variety'' and to ``act like you care about the arguments'' (p.~65). I myself felt, and many colleague debaters also expressed, how the extreme cognitive load that performance in debate demands is incompatible with not fully assuming for the duration of the debate the opinion demanded by the motion, only to then discard it easily in time for the next debate (especially in a competition). Seo corroborates this from the inside: ``Midspeech, a speaker rarely experiences the physical strain of thinking so damn hard. The adrenaline induced a mental fogginess that numbed the hard edges of experience'' (p.~159). This passionate delivery, especially in the context of such extreme cognitive load, is difficult to sustain through meta-cognitive simulation alone. Affective role-assumption becomes the efficient path, organically eliciting rather than faking the emotions expressed.

\textcite{bowman2010functions} identifies nine types of relationships between players and characters, including the ``Fragmented Self,'' where a facet of the player's personality or interests becomes magnified into a central character trait. Debate characters fit this framework: they represent fragmented selves whose defining feature is epistemic commitment to a particular position. The debater role-plays as someone who believes the motion should pass or fail, adopting that argumentative perspective as the character's core identity. They find in themselves beliefs they can hold that support the position and banish from thought all other, focusing on a fragment of their identity.

Second, regarding alibi: this concept describes how game participants can act with lessened social consequences, encouraging risk-taking and willingness to fail \parencite{deterding2017alibis}. Debate provides robust alibi through its competitive frame. Because debate is ``just a game,'' debaters can advance positions they might not advocate publicly and explore uncomfortable arguments. \textcite{seo2022good} captures this vividly: the random assignment of sides provided a paradoxical ``sense of freedom,'' allowing him to ``flirt with ideas, unencumbered by expectations of consistency or deep conviction'' (p.~20). Work simulations do not offer that. This enables the identity experimentation that \textcite{bowman2025transformative} identify as central to transformational containers. It is important to accept that debate however, as all competitive games, does require debaters to perform at and judges them by their peak intellectual ability, not offering them alibi to fail. I would argue however that many TTRPGs and boffer LARPs are designed requiring players to perform intellectually, and that roleplayers within certain communities feel pressure to display acting ability and ability to deeply emotionally immerse. In mainstream North American boffer LARPs, for instance, players who ``break game'' risk derision or ostracism \parencite{brown2016culture}, while in Nordic larp communities, immersion into character is often treated as an indicator of success, with its absence experienced as personal failure \parencite{harder2018larp,pohjola2015steering}. They all however offer players alibi to hold a different identity. Alibi to fail is not essential to a roleplaying game, but alibi to assume the mantle of a character is. On that metric, debate offers players a very strong alibi to embody a character with deeply different opinions from their own: what Seo describes as tracing ``the steps of how a sensible person (us!) could arrive at conclusions that might otherwise have seemed alien'' (p.~224). 

Third, regarding the leisure-community dimension: competitive debate possesses passionate discourse communities distinguishing role-playing games from cousin forms. The World Schools and British Parliamentary circuits encompass hundreds of competitions. Debaters participate voluntarily, often for decades; they develop specialized terminology, shared references, and community norms; they attend tournaments analogous to gaming conventions. The activity is often pursued for its intrinsic satisfactions, not merely as preparation for external goal. This leisure dimension, combined with the vibrant discourse community, positions competitive debate closer to role-playing games than to the educational simulations with which it shares surface similarities.

Furthermore, evidence for debate's role-playing dimension appears throughout its practice. Experienced coaches regularly employ explicit role-playing techniques during preparation, asking debaters to ``become'' stakeholders affected by policies. Furthermore, debate ethos as I witnessed it is summoned through traditions that explicitly stage parliamentary fiction: bench-pounding, heckling, Points of Information delivered as if holding onto wigs. The format performs a fiction of parliamentary deliberation, with debaters and judges alike participating in theatrical conventions inherited from eighteenth-century British debate culture.

Genuine differences between debate and games that are more traditionally seen as analogue role-playing games, such as Dungeons and Dragons, are undeniable: debate lacks the narrative worldbuilding, the character backstories, and the extended campaign structures of tabletop or live-action role-playing. But it is important to recognize that debate's role-playing operates through the same psychological mechanisms and this justifies applying the substantial body of transformative game design scholarship to understand how debate produces its educational effects and, more importantly, how design choices might enhance or undermine those effects.

\subsection{Literature Review of Format Critiques}

The usefulness of applying transformative game design theory to attempt solving the problems of current debate formats comes also from the apparent failure of current scholarly works in the same endeavor. The largest body of work that analyzes debate as an educational tool is that of forensics pedagogy. This centers around the more predominant formats played in the US (especially American Policy). I will thus focus on reviewing what literature was written that echoes the problems I set out to solve and that forms arguments that are also applicable to the formats that I seek to improve (especially the World Schools format).

\subsubsection{The Technical-Public Sphere Tension}

A fundamental critique pervading the literature concerns the evolution of debate formats away from public sphere accessibility toward technical sphere insularity.  \textcite{combs2002psychosocial} documents how both NDT (National Debate Tournament) and CEDA (Cross Examination Debate Association) have evolved into technical sphere activities characterized by: expert-level topic knowledge, speaking rates exceeding what most listeners can process, sophisticated analytical models inaccessible to non-specialists, heavy reliance on extensive note-taking, and specially trained judges.

This has effect both on my critique on value converge as well as related to public speaking abilities.

\subsubsection{Value Convergence and Ideological Homogenization}

A significant concern in the literature relates to what I coined above as value convergence. The ``closed system'' dynamic described by Harris and Smith, cited in \textcite{combs2002psychosocial}, creates conditions conducive to value convergence. In academic debate, tournament practices are insulated from external feedback and where judges who deviate from community norms are ``effectively excluded from the system much like the human system rejects certain bacteria'' \parencite[p.~40]{combs2002psychosocial}. This dynamic may cultivate shared assumptions that, while facilitating intra-community communication, potentially diminish debaters' capacity to engage persuasively with those holding different values.

Crenshaw's work on feminism in debate, cited by \textcite{mcgee2002emancipatory}, illustrates how ``argument borrowing'' can lead to reductionist treatment of complex intellectual traditions. Debaters observe which arguments succeed in out-rounds, then adopt those lines of reasoning for their strategic value rather than out of genuine engagement with the underlying ideas, producing superficial understanding and reduced capacity for authentic cross-ideological dialogue.

\subsubsection{Delivery Practices and Communicative Capacity}

Colbert, cited in \textcite{combs2002psychosocial}, found that average CEDA debaters spoke at approximately 237 words per minute, while NDT finalists averaged 284 words per minute, with some exceeding 300 words per minute, rates that ``greatly exceed what is considered optimal for normal speaking'' \parencite[p.~36]{combs2002psychosocial}.

While World Schools and British Parliamentary formats explicitly discourage talking fast through style-based evaluation, similar pressures toward information density exist. In my observation, this occurs because these formats evaluate on two officially co-equal axes---Style (or Manner) and Content (or Matter)---yet increasing information density improves argumentative quality (more points addressed, more rebuttal, more nuance) while simultaneously reducing intelligibility and persuasive clarity. Because debate operates in an academic setting where the analysis of argumentative quality is perceived as more objective and measurable than the assessment of style, the competitive meta gradually shifts toward privileging Content over Style, even though the evaluation rubric formally treats them as equally important. The result is that the skills developed may not transfer to contexts without expert judges. The difficulty to solve this is also captured in literature. \parencite{hicks2002public} treats the difficulty of defining for what type of audience should debaters make themselves persuasive, comparing \textcite{weiss2002audience}'s ``audience standard'' (being understandable to a regular audience) to his ``ideal of public reason'' (being understandable to the perfect rational listener), presenting how some educational value is lost either way.

\subsubsection{Hierarchical Dynamics and Psychological Safety}

In this issue I have observed forensics pedagogy literature is thinnest, yet several authors recognize the risk. \textcite{snider1999code} addresses safety and ethics in his Code of the Debate, and in his later practical guide warns organizers to ``avoid a counterproductive rigid and hierarchical organization,'' noting ``I've seen groups die when they let ego and status get in the way'' \parencite[p.~36]{snider2014sparking}. The International Debate Education Association's ethics guide for coaches and youth workers similarly acknowledges the ``unbalanced power dynamic'' inherent in coach-student relationships, compounded by the fact that ``older debaters often mentor younger teams'' and thus occupy dual roles of peer and authority figure \parencite{idea2024ethics}. Riley and Hollihan further suggest that approximately three years of debate experience are required to compete effectively on the national circuit, effectively excluding students without extensive preparation and reinforcing experience-based hierarchies \parencite[p.~38]{combs2002psychosocial}.

The mechanism behind these dynamics is illuminated by recent research. \textcite{prussikora2025cracovian}, citing Simon et al.\ (2020), note that ``people placed in competitive intellectual situations tend to develop an adversarial mindset'' characterized by myside bias and otherside bias, distorting cognitive appraisal of the issues being debated. They identify competitive debate's ``most crucial weakness'': its rules ``do not allow for saying to the opponent: `You are right, I was wrong'\,'' instead training debaters to listen for omissions and inaccuracies rather than for understanding \parencite[p.~1474]{prussikora2025cracovian}. When win-loss records become the primary measure of status within these hermetic communities, the adversarial structure produces natural hierarchies in which winning debaters gain social authority while losing debaters are marginalized, reinforcing competitive pressure at the expense of collaborative learning.

\subsection{Gap Identification}

This literature review reveals a significant gap. Forensics pedagogy has established debate's educational benefits and identified tensions, but has not systematically applied game design frameworks to analyze how specific format mechanics produce specific effects. Transformative role-playing game theory provides sophisticated tools for understanding how structured play produces personal change, but has not been applied to competitive debate. Addressing this gap is the central contribution of this thesis.
