\section{Conclusion}

Alfred ``Tuna'' Snider was first to recognize debate as a game and to apply play logic to its analysis, doing this way before game design was a recognized discipline. Through his career as a teacher and scholar he championed the transformative power of playing this game with joy, passion and determination. Although many of his students, one of whom I was very privileged to be, did continue his work in fields such as civics, education, rhetoric and his own specialty of forensics, none used game-design know-how to answer the question that most naturally flows from his argument that debate is a game: how can we make it a better game?

This thesis seeks to begin answering that question. By applying transformative game design theory to competitive debate, it aims to demonstrate that systematic design interventions can address identified problems while preserving debate's documented benefits. The contribution lies not in producing a finished format but in establishing proof of concept: that the intersection of game design scholarship and forensics pedagogy offers productive tools for format improvement, and in articulating design principles worthy of further development and investigation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TIMELINE ANNEX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Timeline Annex}
\addcontentsline{toc}{section}{Timeline Annex}

\begin{itemize}
    \item \textbf{01--15 January:} Thesis proposal refining
    \item \textbf{16--31 January:} Design of the new format, conversations with fellow senior debaters and with other game designers; organizing the first playtest
    \item \textbf{01--15 February:} First pilot playtest, first interviews, announcing the mini-competition and opening registration, iteration on the game design
    \item \textbf{16--28 February:} Second pilot playtest and interviews, finishing registrations for the mini-competition, iteration on the game design
    \item \textbf{01--20 March:} Mini-competition extensive playtests, running all remaining interviews
    \item \textbf{21 March--01 May:} Data analysis, further theoretical inquiries, thesis writing
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DEBATE FORMAT ANNEX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Annex: Inside Out Debate Format}
\addcontentsline{toc}{section}{Annex: Inside Out Debate Format}

\subsection*{Team Structure}

Debaters are organised in four teams of two debaters each and are assigned positions within the debate at random:

\begin{itemize}
    \item \textbf{Team S (Subject):} Represents the Subject of the Motion, a person who has to take a decision.
    \item \textbf{Teams A, B, and C (Value Defenders):} Each must argue that a particular Value held by the Subject should be prioritised in the decision, and consequently, what the Subject should decide.
\end{itemize}

\subsection*{The Motion}

The Motion outlines:

\begin{itemize}
    \item What is the decision that has to be made.
    \item What is the background of the Subject and other relevant practical factors that can be taken into account within the decision.
    \item What are the Values of the Subject.
\end{itemize}

It is important to note that the Motion does not prescribe to the Value Defenders a particular position on what decision the Subject has to make. They have to construct it themselves.

\subsection*{Speech Order}

The four teams hold alternating speeches in the following order:

\begin{center}
A1, B1, C1, S1, A2, B2, C2, S2
\end{center}

Within their speeches, the Value Defenders can and should construct arguments in favour of both the relevance of their value and the particular decision they constructed, as well as engage with the arguments forwarded by the other Value Defenders.

Within their speeches, the Subject team must outline a strong reasoning that engages with and explains the interplay between all arguments of the Value Defenders, picking a particular decision. Within S1 they pick a preliminary decision and within S2 a final decision.

\subsection*{Evaluation}

The debate is overseen by an Evaluator, who performs two tasks:

\paragraph{Commendations.} The Evaluator awards Commendations to debaters. Commendations represent acknowledgements of useful skills, knowledges, or attitudes. A list of awardable commendations is constructed by the Chief Evaluators of the Debate Tournament, along with the total number of Commendations that Evaluators can offer within that Tournament. Example commendations include:

\begin{itemize}
    \item Compassionate Discourse
    \item Intriguing Argumentation
    \item Strategic Rebuttal
    \item Inspiring Team Leadership
\end{itemize}

Over time, debaters who gather a particular commendation at least three times earn a Badge. Gathering it additional times upgrades the badge, an unlimited number of times, but each upgrade requires gathering three more Commendations than last time.

\paragraph{Oral Feedback.} The Evaluator delivers an Oral Feedback session, in which they explain to each team the scores and commendations they received, the reasoning behind them, and give advice on how each debater can improve in the future based on their performance.

\subsection*{Example Motion}

\begin{quote}
You are the parent of a 16-year-old teenager. Your family is middle class and lives in the suburbia of a metropolis. You have higher education. You live in a country in which Higher Education is free, but there are only a few universities that are deemed as good. The year is 2026. You evaluate that your kid is talented in music, is a good leader and has had very good STEM skills, but is also quite distracted, does not have many friends and has failed to learn a foreign language. Your kid expressed a moderate interest in studying Musicology at university after they finish high school, but is also considering not going to higher education. You have a good relationship with your child. They asked you to suggest what they should do. What do you recommend your kid they should do after they finish high school?

\medskip

Three Beliefs that you hold, relevant for this decision, are:
\begin{itemize}
    \item In life, financial stability and independence is essential for a happy life.
    \item It is important to be a well-rounded, cultured individual.
    \item The people that you surround yourself with end up defining who you are.
\end{itemize}
\end{quote}
