\section{DESIGN OF THE INSIDE OUT FORMAT}

\subsection{From Theory to Design}

This chapter corresponds to the generation and critical scrutiny phases of the Research through Design methodology outlined in the Methods chapter. The theoretical framework produced three design predictions constituting the design brief: formats maximizing transformation should (1) structure memetic bleed toward epistemic humility, (2) align procedural bleed with transferable communication skills, and (3) maintain container safety by preventing competitive success from conflating with personal worth. The following section presents the Inside Out format as a designed artifact, before the remainder of the chapter argues why each element exists.

\subsection{The Inside Out Debate Format}

\subsubsection{Team Structure}

Debaters are organised in four teams of two debaters each and are assigned positions within the debate at random:

\begin{itemize}
    \item \textbf{Team S (Subject):} Represents the Subject of the Motion, a person who has to take a decision.
    \item \textbf{Teams A, B, and C (Value Defenders):} Each must argue that a particular Value held by the Subject should be prioritised in the decision, and consequently, what the Subject should decide.
\end{itemize}

\subsubsection{The Motion}

The Motion outlines:

\begin{itemize}
    \item What is the decision that has to be made.
    \item What is the background of the Subject and other relevant practical factors that can be taken into account within the decision.
    \item What are the Values of the Subject.
\end{itemize}

It is important to note that the Motion does not prescribe to the Value Defenders a particular position on what decision the Subject has to make. They have to construct it themselves.

\subsubsection{Speech Order}

The four teams hold alternating speeches in the following order:

\begin{center}
A1, B1, C1, S1, A2, B2, C2, S2
\end{center}

Within their speeches, the Value Defenders can and should construct arguments in favour of both the relevance of their value and the particular decision they constructed, as well as engage with the arguments forwarded by the other Value Defenders.

Within their speeches, the Subject team must outline a strong reasoning that engages with and explains the interplay between all arguments of the Value Defenders, picking a particular decision. Within S1 they pick a preliminary decision and within S2 a final decision.

\subsubsection{Evaluation}

The debate is overseen by an Evaluator, who performs two tasks:

\paragraph{Commendations.} The Evaluator awards Commendations to debaters. Commendations represent acknowledgements of useful skills, knowledges, or attitudes. A list of awardable commendations is constructed by the Chief Evaluators of the Debate Tournament, along with the total number of Commendations that Evaluators can offer within that Tournament. Example commendations include:

\begin{itemize}
    \item Compassionate Discourse
    \item Intriguing Argumentation
    \item Strategic Rebuttal
    \item Inspiring Team Leadership
\end{itemize}

Over time, debaters who gather a particular commendation at least three times earn a Badge. Gathering it additional times upgrades the badge, an unlimited number of times, but each upgrade requires gathering three more Commendations than last time.

\paragraph{Oral Feedback.} The Evaluator delivers an Oral Feedback session, in which they explain to each team the commendations they received, the reasoning behind them, and give advice on how each debater can improve in the future based on their performance.

\subsubsection{Example Motion}

\begin{quote}
You are the parent of a 16-year-old teenager. Your family is middle class and lives in the suburbia of a metropolis. You have higher education. You live in a country in which Higher Education is free, but there are only a few universities that are deemed as good. The year is 2026. You evaluate that your kid is talented in music, is a good leader and has had very good STEM skills, but is also quite distracted, does not have many friends and has failed to learn a foreign language. Your kid expressed a moderate interest in studying Musicology at university after they finish high school, but is also considering not going to higher education. You have a good relationship with your child. They asked you to suggest what they should do. What do you recommend your kid they should do after they finish high school?

\medskip

Three Beliefs that you hold, relevant for this decision, are:
\begin{itemize}
    \item In life, financial stability and independence is essential for a happy life.
    \item It is important to be a well-rounded, cultured individual.
    \item The people that you surround yourself with end up defining who you are.
\end{itemize}
\end{quote}

\subsection{Design Rationale}

Each structural choice in Inside Out traces to the theoretical framework's diagnosis of how traditional formats produce corrupted bleed, misaligned skill transfer, and container compromise. This section justifies the format's core design decisions.

\subsubsection{Why Multiple Teams Defending Values?}

Traditional formats position two teams in binary opposition, a structure whose procedural rhetoric \parencite{bogost2007persuasive} teaches that every issue has exactly two sides. Inside Out replaces this with three Value Defender teams, each advocating for a different value held by the Subject. This restructuring addresses several identified problems simultaneously.

The multi-team structure produces a procedural rhetoric of complexity and humility: the format asserts through its rules that decisions involve balancing multiple legitimate human values, not choosing between right and wrong. This directly counters the corrupted memetic bleed driving value convergence by positioning value pluralism at the core of the conflict rather than treating it as a side effect of switch-side argumentation. Because values are assigned rather than chosen, the design also strengthens alibi \parencite{deterding2017alibis} for assuming a fragmented identity \parencite{bowman2010functions}, the debater-character relationship is explicitly about inhabiting a value, making perspective-taking the core mechanic. And because the Subject is a person with specific background, circumstances, and values rather than an abstract decision-maker, the motion reinforces that decisions are products of context, resisting the abstraction that enables generic argument templates and the drift toward what \textcite{goodnight1982spheres} identified as the technical sphere of argumentation.

Furthermore, the structure cultivates collaborative alongside combative skills. Value Defenders must persuade the Subject, a sympathetic listener making a genuine decision, not defeat an opponent. This corrects the misaligned procedural bleed by developing communication habits adapted to contexts beyond competitive debate, where persuasion typically involves collaboration rather than combat \parencite{cirlin1997sociological}. The Subject also creates a natural accessibility incentive: if a Value Defender speaks incomprehensibly, their arguments are simply not weighed in the decision, a consequence that recalibrates the zone of proximal development \parencite{vygotsky1978mind} toward general-audience communication without any external rule imposing it.

\subsubsection{Why Split Subject and Evaluator?}

In traditional formats, a single judge both determines the winner and provides feedback, entangling evaluation with adjudication. Inside Out separates these functions: the Subject makes decisions within the debate, while the Evaluator observes and offers commendations.

This split means winning the debate is not a goal in itself but a method for growing, symbolized through commendations rather than victories. Decision-making within the debate becomes an act not of arbitrating but of synthesizing toward a position by acknowledging the interplay between values, a form of procedural rhetoric that models how real-world deliberation works, steering corrupted memetic bleed away from value convergence and toward epistemic humility. Moreover, the Subject's defined character invites bleed-in from character to debater, making the player less receptive to unpersuasive high-density argumentation precisely because the character they inhabit would not be. The Subject learns to synthesize; the Value Defenders learn to persuade collaboratively; and the Evaluator, freed from adjudicating competition, can focus on recognizing communicative virtue, maintaining the transformational container by decoupling status from competitive victory.

\subsubsection{Why Commendations Instead of Scores?}

Replacing scores and victory with commendations moves the format from a paradigm of competing against others toward one of improving oneself. The commendation system recognizes individual worth: the aspiration is not to become the best debater, but to become the best form of debater one can become. Two debaters can be excellent in entirely different ways, a message reinforced by the diversity of commendation categories (``Compassionate Discourse'' alongside ``Strategic Rebuttal'') and the badge progression system that builds unique skill profiles over time. This directly addresses the container compromise identified in the theoretical framework: when status derives from demonstrated growth rather than competitive victory, the zero-sum dynamics that erode the transformational container \parencite{bowman2021magic, bowman2025transformative} are structurally dismantled.

The commendation system also resolves a persistent tension in modern debate formats between content density and persuasive discourse, precisely the misaligned procedural bleed identified in the theoretical framework. Commendations dissolve this conflict: debaters can pursue recognition for persuasive discourse alongside and equally with recognition for academic rigor and complexity, because these are separate categories rather than competing dimensions collapsed onto a single scale.

\subsubsection{Why Oral Feedback Instead of Oral Critique?}

In traditional formats, the oral critique centers on the judge's decision: the judge explains who won and why, and advises debaters on what would have changed the result. The implicit question is ``how could I have won?'' Inside Out reframes post-round interaction as oral feedback, centering on demonstrated skills and developmental trajectory. The Evaluator explains the commendations awarded, articulates why particular moments demonstrated particular skills, and offers advice oriented toward growth. The implicit question shifts to ``what am I developing, and how can I develop further?'' This connects the debate experience to the debater's broader growth trajectory rather than to a single round's competitive outcome, maintaining the transformational container by structuring the debrief around growth rather than victory.

I acknowledge that the current design does not yet include explicit emotional processing or de-roling practices, mechanisms that \textcite{bowman2025transformative} identify as important for transformative game design. This represents an area for iteration in future design cycles.

\subsection{Synthesis and Design Tensions}

Taken together, the design decisions map onto the three identified problems as follows. Corrupted memetic bleed (value convergence) is addressed by the multi-team structure, which positions value pluralism at the core of the conflict; by the personal Subject, which grounds debate in context-dependent reasoning; and by commendations, which sever the evaluative feedback loop. Misaligned procedural bleed (information density) is addressed by the Subject as accessibility anchor, by the collaborative persuasion dynamic, and by commendation categories that reward persuasive discourse alongside analytical rigor. Container compromise (competitive pressure) is addressed by commendations replacing victory, by the Evaluator's decoupled role, and by oral feedback oriented toward growth. These interactions suggest that the three problems are not independent pathologies but interconnected symptoms of deeper structural features in traditional formats; Inside Out's design elements form a coherent alternative system addressing them jointly.

Several tensions in this design constitute productive questions for playtesting. Will competitive motivation survive without win-loss outcomes, or will debaters socialized in traditional formats experience commendations as insufficient stakes? The Subject team bears a unique cognitive burden in synthesizing three value streams, and whether this is productive challenge or overwhelming difficulty depends on experience level and motion complexity. The format's novelty (four teams, new roles, unfamiliar mechanics) may create barriers to entry for newcomers. Without victory as evaluative anchor, commendation judgments may feel more subjective; clear criteria and Evaluator training will likely be essential. Finally, the format does not yet include explicit emotional processing or de-roling practices, and whether reduced competitive pressure makes these less necessary remains an open question.
