\section{Discussion}

\subsection{Debate as a Theoretically Productive Test Case}

Transformative game design theory has been developed primarily through the study of live-action role-playing and tabletop role-playing games \parencite{bowman2025transformative}. In these contexts, alibi is informal and socially negotiated rather than institutionally enforced; evaluation is absent or peer-governed rather than adjudicated through explicit criteria; and community membership is largely voluntary, low-stakes, and self-selected. These features have shaped the theory's assumptions in ways that rarely become visible within the contexts where the theory was built, because the contexts do not stress-test them.

Competitive debate differs structurally in three respects that function as theoretical amplifiers. First, its alibi is institutionalized: side assignment is random, publicly documented, and enforced by the format's rules, not merely understood as a social convention. Debaters cannot choose which position they argue any more than they can choose their speaking slot, which means the alibi is available regardless of the participant's disposition to use it. Second, debate has a formalized evaluation tradition with historical precedent: oral critique, speaker points, round records, and win-loss tabulation provide a developed apparatus for assessing what happens during play, an apparatus that most transformative play contexts entirely lack. Third, debate operates within pre-existing communities with defined status hierarchies, competitive socialization, and accumulated habitus---communities that arrive at a new format with expectations, investments, and resistances already in place. These three features do not make debate a better or worse context for transformation than LARP or tabletop play. They make it a different one, and that difference is what makes it theoretically productive.

What follows is not a report on whether Inside Out works---the Results chapter addressed that question through the proof-of-concept framework---but an account of what the encounter between this design and these participants reveals about the theory that generated it. Each of the observations discussed below arose from a specific empirical surprise: a moment where the data did not behave as the design predicted, or where it confirmed a prediction in an unexpected way, or where a mechanism operated through a pathway the theory had not anticipated. These surprises are not design failures. They are the conditions under which theory becomes visible to itself.

\subsection{The Alibi Paradox: Role Depth Channels Bleed Rather Than Buffering It}

The Inside Out format assigns participants to roles of two distinct types: the three Inner Voice teams, each defending one of the Subject's values, and the Subject, who inhabits a specific named character with defined relationships, background, and stated tensions. The Subject role was designed to carry substantial psychological texture precisely because that texture was expected to strengthen the alibi. If the Subject is not an abstract decision-maker but a fully specified individual---Mara, a teacher, a parent, facing a particular dilemma---then the player's assignment to the role is unambiguous, and the deflection of inference that \textcite{deterding2017alibis} identifies as alibi's core function should be correspondingly robust. The player argues from the character's position, not their own.

What the playtests revealed was the opposite. The Subject's opening speech---designed to expose the character's values, biases, and emotional stakes so that the Inner Voice teams could calibrate their persuasion---caused the Subject's own biases to bleed into the character rather than the character's biases to provide cover for the player. In the second playtest, one participant's opening monologue was noted to contain ``nuanced biases that bled from the player to the character,'' and Inner Voice teams immediately began strategizing around those personal biases rather than around the character's stated profile. The distinction between player and character that the detailed character design was meant to establish collapsed precisely where the design had worked hardest to enforce it. The most psychologically elaborate role produced the most intense identification, not the most reliable separation.

This observation is consistent with the neuroscientific account of bleed that \textcite{hugaas2024bleed} draw upon: embodied cognition means that inhabiting a psychologically rich role activates neuropsychological processes that do not readily distinguish the fictional from the real. A shallow role assignment---``you are the Proposition''---engages the player's argumentation without deeply engaging their identity. A detailed character with specified values, relationships, and emotional stakes engages the player's identity precisely, pulling from fragments of the self that the character's design happens to resonate with. The richer the role, the more genuine the identification, and the more genuinely the player's own perspective enters the character rather than remaining safely behind it.

This suggests that \textcite{deterding2017alibis}'s alibi concept needs to be disaggregated. The existing formulation treats alibi as a single mechanism whose strength is a function of how clearly a role is assigned---the cleaner the assignment, the more robust the deflection. But the data indicates that alibi operates through at least two qualitatively distinct functions rather than a single spectrum of strength. The first is \textit{alibi-as-buffer}: a shallow role assignment provides a social excuse that deflects inference without generating deep identification. The debater arguing for a side they were randomly assigned has alibi in this sense---they can say ``I was assigned this,'' and both they and their audience understand that the argument does not reflect their conviction. Bleed under buffer alibi is minimal, because the role has not engaged the player's identity deeply enough to activate the biographical material that bleed draws upon.

The second function is \textit{alibi-as-channel}: a psychologically rich role assignment generates imaginative identification that activates biographical material and directs it into the character, producing intense bleed along a pathway shaped by the character's design. The Subject in Inside Out has alibi in the sense that they did not choose their situation---it was assigned to them---but the character's psychological specificity means the assignment engages the player's own values, experiences, and emotional responses in a way that a simpler role would not. The alibi provides the social license to inhabit the role without social risk; the depth of the role determines what gets activated once the player accepts that license.

Buffer alibi and channel alibi are not points on a single dimension of alibi ``strength.'' They are different mechanisms serving different purposes. Buffer alibi protects by creating distance; channel alibi directs by creating proximity. Designers must choose which function a given role needs to serve, because deeper character work does not produce a safer container---it produces a more potent one. The Inside Out Subject role was designed to function as a buffer while its psychological texture was already producing channel effects, a productive misfire that makes the distinction legible in a way it had not been in the existing literature.

The strongest bleed event in the dataset---a participant privately disclosing, during a break between rounds, that the debate had mirrored a real family situation encountered that morning---occurred in the Subject role. This was spontaneous, private, and behaviorally unprompted: precisely the signature of genuine bleed rather than performed response to social expectation. It is significant that this event arose not from the Inner Voice roles, which carry lighter psychological texture, but from the role designed to carry the most. Deep alibi had channeled the participant's real life into the game, and then back out again.

\subsection{What Rules Reward, Not What They Encode: Procedural Rhetoric Revised}

The Inside Out format's three-team Inner Voice structure was designed to encode a specific worldview through its rules: that significant decisions involve the simultaneous pull of multiple legitimate human values, none of which simply overrides the others, and that the task of deliberation is to navigate that irreducible complexity rather than to identify the correct answer. In \textcite{bogost2007persuasive}'s terms, the format was designed to make a procedural argument for value pluralism. The argument was to be made not through propositional statement but through the format's logic itself: by requiring participants to defend a single value among peers defending others, and by requiring the Subject to weigh all three, the format would model the structure of genuine moral dilemmas.

What the second playtest revealed was a systematic drift away from this. By the second round, play had shifted from defending distinct values toward arguing for a ``dominant lens''---a framing through which all of the Subject's values could be synthesized, and which the Subject should adopt as the organizing principle for their decision. Rather than maintaining the tension between irreducibly different values, Inner Voice teams began converging on a shared meta-level argument about which value's logic should take precedence. The designer's own observation, playing as an Inner Voice participant, was that ``arguing for my lens and shifting focus felt like a more natural and persuasive strategy, whilst accepting points made by fellow Value Defenders, or even allowing myself to be persuaded to adjust the team's stance based on the points made by them.'' This is not value pluralism. It is collaborative meaning-making oriented toward synthesis.

The mechanism is not difficult to identify once the incentive structure is examined. The format requires the Subject to make a decision. Whatever the three Inner Voice teams argue, the Subject must ultimately arrive at a position---must synthesize the three value streams into a coherent basis for action. Inner Voice teams, aware of this requirement, rationally orient their arguments toward facilitating that synthesis in their favor. The path of least resistance is not to insist on the incommensurability of one's own value with the others, but to demonstrate that one's value provides the most useful organizing framework for the Subject's decision-making. Convergence toward a dominant lens is not a failure of the design's pluralist intent; it is the rational response to the format's incentive logic.

This exposes a gap in the existing formulation of procedural rhetoric. \textcite{bogost2007persuasive} argues that games make arguments through their rules and systems, and that players internalize the worldviews those systems embody through the act of play. But the Inside Out case suggests that there are two distinct levels at which a game's system can ``speak,'' and they can diverge. The first is the \textit{encoded worldview}: what the rules represent, symbolize, or aspire to teach. Inside Out's three-team structure encodes value pluralism at this level---three distinct values are treated as simultaneously worthy of defense, and the format's surface logic asserts their equality. The second is the \textit{incentivized worldview}: what the rules reward optimizing for. Because the Subject must decide, and because Inner Voice teams know this, the incentivized logic is convergence toward synthesis. This is what the format actually teaches through play.

When encoded and incentivized worldviews diverge, the incentivized worldview is the actual procedural argument being made by the format. Players respond to incentives, not representations. A format that symbolically encodes pluralism while incentivizing synthesis is, procedurally, teaching synthesis. For transformative game design, this distinction carries a practical implication: structural symbolism is insufficient to guarantee that a format teaches what it means to teach. Designers must audit not only what their rules represent but what their rules reward, because it is the latter that operates on participants through the logic of play.

The same analysis offers a more positive reading of one of the results chapter's strongest findings. The emergent norm against adversarial rebuttal---a community-level behavioral pattern that arose within a single session without being specified in the ruleset---is procedural rhetoric operating as intended. The format rewards collaborative persuasion: Inner Voice teams are not trying to defeat each other but to help the Subject decide well, and adversarial rebuttal interferes with that goal. Participants internalized this incentive logic rapidly and expressed it as a felt norm rather than a strategic calculation. This is a significant observation in its own right, because it suggests that procedural rhetoric operates not only at the level of individual cognitive internalization, which was \citeauthor{bogost2007persuasive}'s original framing, but also at the level of emergent community norms formed within the space of a single session. The format reshaped participants' shared understanding of what appropriate conduct looks like, and it did so through incentive logic rather than instruction.

\subsection{Container Robustness and the Two Levels of Psychological Safety}

\textcite{bowman2025transformative} caution that a transformational container's effectiveness is not guaranteed by its mere existence, and that community support is necessary to sustain the conditions under which transformation occurs. The playtests produced what appeared at first to be a contradiction of this principle: psychological safety was established immediately, robustly, and consistently across both sessions, despite the absence of any mature community structure, any shared history between participants and the format, and any established norms for how the container ought to function. Participants who had never encountered Inside Out reported dramatically reduced anxiety relative to their experience in traditional formats, and none of the predicted negative indicators of container compromise---distress, arrogant behavior, adversarial posturing---manifested across either session. The container appeared to be structurally robust in a way the theoretical caveat had suggested it might not be.

And yet the same playtests revealed that evaluators found commendation-giving genuinely difficult. One evaluator reported that it was ``very difficult to find feedback to give, because the format does not specify so specifically towards what educational objectives should the feedback be given.'' Another found commendation assignments hard without the reference point of a larger circuit where relative achievement would be more visible. The container was established---participants felt safe, engaged willingly, took risks, and showed evidence of genuine perspective work. But the evaluative apparatus designed to reinforce and direct transformation could not fully perform its function, because what growth looked like in this format had not been made sufficiently legible.

These two observations point toward a distinction the existing theory does not make explicitly. Container function appears to operate at two levels, which are structurally different and require different design attention.

The first level is \textit{container establishment}: the experience of reduced identity threat, diminished anxiety, and willingness to engage without defensive self-protection. The data suggests this level is achievable through format design alone, without requiring mature community infrastructure. Removing win-loss outcomes as the primary evaluative metric, assigning roles that provide explicit alibi, and contextualizing the motion around a specific person's values rather than an abstract proposition all contributed to immediate safety. These are structural features, and their effect was rapid---present in the first session with participants encountering the format for the first time. The container's establishment is, in this sense, a more tractable design problem than the theory's caution implies.

The second level is \textit{container utilization}: the capacity of the established container to produce, reinforce, and consolidate the specific transformative outcomes the format is designed to generate. This level requires more than structural safety. It requires that what transformation looks like be made legible---to participants, so they can recognize when it is happening; to evaluators, so they can notice and name it in the debrief; and to the format itself, so the evaluative apparatus can connect the experience to participants' broader development. Where this legibility is absent, the container is established but underutilized: participants feel safe and engage genuinely, but the experience does not reliably integrate into lasting change because the apparatus designed to facilitate that integration cannot perform its function.

The thesis's commendation problem is the gap between these two levels made concrete. The container was established: safety was present, engagement was genuine, and bleed occurred. But without legible growth criteria, the debrief could not connect specific moments of bleed or perspective change to the participants' ongoing development trajectories. Transformation happened---the spontaneous bleed event in the second playtest is the clearest evidence of this---but it happened despite the underspecified utilization level, not because of it. A more deliberately designed second level would have provided participants and evaluators with the vocabulary to recognize what had occurred and to use it.

The theoretical implication is that \textcite{bowman2025transformative}'s caution about community support should be disaggregated into two separate claims. Community support is primarily a condition for the second level of container function---for utilization rather than establishment. Establishment can be achieved structurally and relatively quickly. The risk is that researchers and designers mistake the establishment of container safety, which is measurable and immediate, for the utilization of that safety toward lasting transformation, which is slower, more internal, and harder to observe. A format that produces positive experiences without legible growth criteria may establish the container without effectively using it.

\subsection{Emancipatory Bleed and the Prior Status Problem}

\textcite{kemper2017battle,kemper2020wyrding} describe emancipatory bleed as a mechanism through which marginalized participants, given a container in which they are taken seriously on equal terms, experience lasting changes in self-concept and capability. The background chapter documents substantial empirical evidence for this effect in competitive debate: Urban Debate Leagues and similar programs have produced documented outcomes for previously ``silenced student populations'' \parencite{snider2003gamemaster}. The theoretical framework treats this as a mechanism whose availability scales with the quality of the container: design a better container, and emancipatory bleed becomes more reliably available across the participant population.

The playtests introduced a complication the theory did not anticipate. Responses to Inside Out were not uniform across participants but showed a consistent pattern of differentiation correlated with participants' prior experience of the existing competitive system. Participants who had been less successful in traditional formats, and women across both groups, consistently reported strong preference for the new format, describing reduced anxiety, reduced ego-investment in outcomes, and a sense of liberation from the evaluative pressure of win-loss competition. Participants who had been highly successful in traditional formats expressed a different response: they found the format less exhilarating, and articulated specifically what it removed. As one participant put it, ``there are few things more ego-boosting than the feeling of total mind-domination that you get when you win a debate, and this cannot give you that.'' No highly successful competitor expressed enthusiasm for the format's transformative potential; all the enthusiasm came from participants for whom traditional formats had been more costly.

This pattern is not incidental. Emancipatory bleed operates by providing a container in which participants who have been marginalized by the existing system can be taken seriously on equal terms. Its availability is therefore highest for participants who have the most identity stake in that equalization---those for whom the existing system's evaluative mechanisms have most consistently registered their contributions as insufficient, unconvincing, or lower-status. For participants who have already secured recognition through those same mechanisms, the format's redesign removes something they have invested in and rewarded. The container equalization that liberates one participant is experienced by another as the removal of a meaningful distinction they earned.

This is what might be called the \textit{prior status problem} in transformative game design. When a transformative format is applied not to a new community beginning without history but to an existing community with established status hierarchies and competitive socialization, the experience of transformation is not uniform. It is liberation for those the existing system marginalizes, and diminishment for those it rewards. This is not a bug in the design. It is a structural feature of applying transformation to communities rather than to individuals, and the existing theory does not account for it.

The prior status problem carries three implications for transformative game design as a field.

First, transformation readiness is unevenly distributed within existing communities, and its distribution is sociological rather than psychological. It is not that some participants are more open-minded or more resistant to change as individuals. It is that their relationship to transformation is shaped by their accumulated position within the status architecture of the community the format is asking them to change. Designing for transformation in an existing community therefore requires mapping who stands to gain from the change and who stands to lose it, not as a question of individual temperament but as a question of social position. \textcite{bowman2025transformative}'s list of designable surfaces does not include the community's existing status architecture, but the data suggests it should.

Second, format redesign is necessary but not sufficient for community transformation. Inside Out can redesign the evaluative apparatus, the role structure, and the feedback framework---all of which it does---without directly addressing the cultural infrastructure of the community surrounding it: the norms about what success means, who is centered, what behaviors are admired, and what is expected of participants at different levels of experience. A format introduced into a community carries the community with it. Participants bring their competitive habitus to the new format and apply it to whatever evaluative handles are available. Format change and community cultural change must be addressed in parallel rather than sequentially.

Third, the population most resistant to transformation within an existing debate community is also the population whose behavior most reproduces the problems the format is designed to address. Value convergence, information density, and hierarchical dynamics are not produced by participants who feel marginalized by the existing system; they are produced by participants who have been rewarded by it and who have most thoroughly internalized its logic. These participants are the primary target of the transformation Inside Out is designed to catalyze, and the data shows they are the least receptive to it. This is not a problem that format design can resolve alone, and the field should be explicit about that limitation rather than implying that a well-designed format will produce transformation uniformly across any community it encounters.

\subsection{Growth Legibility: The Undertheorized Operational Challenge}

The commendation system was designed to address the most structurally damaging feature of traditional debate evaluation: the zero-sum character of win-loss outcomes, in which one team's gain is definitionally another's loss, and in which competitive success becomes the primary currency of status within the community. By replacing scores and victory with commendations---individualized recognitions of specific demonstrated skills, awarded independently to each participant---the system was designed to decouple status from competition, orient the debrief toward growth rather than adjudication, and create a progression structure that accumulates over time rather than resetting with each round. The theory predicted that this redesign would maintain the transformational container's function while removing the mechanism most responsible for its compromise.

The playtests confirmed the decoupling and demonstrated genuine enthusiasm for the progression logic---one participant described the commendation system as scratching ``the 'make a build' itch,'' and the language of leveling up and developing a skill profile was adopted readily and with evident pleasure. But evaluators found the system operationally difficult in ways the theory did not anticipate. One evaluator reported that commendation-giving was difficult without the reference points provided by a larger circuit where relative achievement would be more visible, and found it hard to specify ``what educational objectives should the feedback be given.'' Another noted that commendation criteria risked ``farming''---participants choosing stylized or eccentric strategies specifically to accumulate commendations in underrepresented categories, rather than pursuing genuine development.

Both difficulties point to the same underlying problem, which is that the commendation system decoupled status from competition without solving the problem of evaluation. Win-loss evaluation is legible: it is binary, externally observable, and requires no interpretation beyond identifying which team persuaded the judge. The commendation system replaced this legibility with something that has no obvious equivalent. What does ``demonstrated epistemic humility'' look like in a two-minute speech? How does an evaluator distinguish genuine perspective change from rhetorically performed openness? When is a moment of synthesis in the Subject's closing speech evidence of transformative engagement, and when is it good debate craft? These questions did not arise in traditional debate because traditional debate did not ask evaluators to assess transformation. Inside Out asks evaluators to do exactly this, without yet providing the frameworks that would make such assessment possible.

This surfaces \textit{growth legibility} as a fundamental and undertheorized design challenge in transformative game design. The challenge has two distinct components that require separate design solutions.

The first is \textit{observational legibility}: the capacity of a facilitator to recognize growth as it occurs. Observational legibility requires at minimum three things. It requires prior specification of what growth looks like in behavioral terms, not just conceptual terms: not ``epistemic humility'' as an aspiration but ``participant adjusts stated position in response to another's argument without defensive framing'' as an observable event. It requires facilitators who can distinguish genuine transformation from performance of transformation---a distinction that social desirability effects make systematically difficult, especially in first-contact sessions where participants are motivated to demonstrate engagement. And it requires evaluation instruments calibrated to the specific transformative goals of the particular format, rather than generic skill categories that apply equally to any competent public speaker.

The second component is \textit{anti-farming legibility}: the capacity of the progression system to be designed such that optimizing for commendations is equivalent to pursuing genuine growth. Any progression system---competitive or not---will attract optimization behavior. This is not a pathology of competitive debaters but a rational response to any system that offers designable rewards for designable behaviors. The design question is not whether to optimize but what to make the target of optimization. If commendation criteria are well-specified and genuinely track the transformative outcomes the format is designed to produce, then farming commendations \textit{is} developing the corresponding capacities, and the optimization pressure serves rather than undermines the format's goals. If criteria are underspecified, farming produces hollow performances---the appearance of the desired behavior without the underlying development---and the progression system actively interferes with transformation by creating incentives to simulate it.

These two legibility challenges are not specific to Inside Out. They are present in every transformative game that replaces external competitive evaluation with growth-oriented progression. The reason they have not been theorized by the field is that most transformative LARP and RPG research does not include formalized evaluation systems at all: participants self-assess, peers offer informal feedback, and the question of what growth ``looks like'' to an evaluator does not arise because there is no evaluator in the relevant sense. Competitive debate, precisely because it has an institutionalized evaluation tradition with trained adjudicators and established feedback conventions, makes the legibility gap visible in a way that softer play contexts entirely obscure. The gap does not appear in those contexts because it is never required to fill---not because those contexts have solved it. Inside Out, by introducing growth-oriented evaluation into a tradition that has always had evaluation, makes it impossible to avoid.

Growth legibility should therefore be understood as a designable surface in its own right---one that requires as much deliberate attention as role structure, motion design, or feedback timing. The field's existing accounts of what makes a transformational container effective do not yet include it, and its absence from the list is not a minor gap. An evaluative apparatus that cannot reliably recognize transformation cannot reliably reinforce it; and transformation that is not reinforced through the debrief---which \textcite{bowman2025transformative} identify as the primary site where raw experience is converted into lasting change---remains a fortunate accident rather than a designed outcome.

\subsection{Synthesis: The Field's New Picture}

Taken together, the five sections of this discussion articulate what is revealed when transformative game design theory encounters competitive debate's institutional structure. The encounter produces not a verdict on the theory's validity---the proof-of-concept results demonstrate that the framework is productive---but a set of extensions and disaggregations that the theory requires in order to operate at the level of precision that intentional design demands. It is worth noting, however, that several of the most striking empirical outcomes emerged not because the theory correctly predicted them but in spite of its predictions: the spontaneous bleed events, the emergent collaborative norms, and the immediate container establishment all arose through pathways the theory had not anticipated. This does not diminish the framework's value, but it suggests that the relationship between transformative game design theory and practice is generative rather than deductive---the theory provides a productive starting point, not a reliable blueprint, and its value is measured by what it makes visible when it fails as much as by what it predicts when it succeeds.

Four underspecifications have been identified. Alibi, treated in the existing literature as a single mechanism whose strength varies along a spectrum from weak to robust, is better understood as two qualitatively distinct functions: buffer alibi, which creates distance, and channel alibi, which creates directed proximity. Depth of character design determines which function a role serves, and designers must choose explicitly rather than assuming that greater role specificity produces greater protection. Procedural rhetoric, understood as the internalization of worldviews encoded in rules, needs to distinguish between what rules represent and what rules reward, because these can diverge significantly; the actual procedural argument made by a format is the incentivized worldview, not the encoded one. The transformational container, understood as a binary that is either established or compromised, operates at two levels---establishment and utilization---that require different design attention and that can come apart; a format can successfully establish psychological safety without creating the conditions for that safety to produce lasting change. Growth legibility, not yet theorized in the existing framework, is the operational prerequisite for the debrief to perform its transformative function, and without it any evaluation system that replaces competition with progression will face both the observational problem of recognizing genuine transformation and the optimization problem of designing progression that cannot be gamed away from its intended outcomes.

The prior status problem identified in the discussion of emancipatory bleed is of a different character. It is not an underspecification within the theory's existing scope but an extension of that scope. The theory has been developed in contexts where transformative game design is introduced to communities that begin without strong prior status architecture---or where the researcher's view does not foreground that architecture. Applied to existing communities with deeply ingrained competitive habitus, the theory reveals that format transformation and community transformation are different problems, that they require different interventions, and that the relationship between them is not sequential but structural. Format redesign can create the conditions for a different kind of community to emerge; it cannot, by itself, transform the community that arrives at the new format carrying the old one's values.

What this thesis does not claim is as important as what it does. The playtests involved small numbers of experienced participants in a single cultural context over a short timeframe. They provide no evidence about longitudinal effects, about the format's function with novice debaters, about its transferability across linguistic and cultural communities, or about whether the anxiety reductions and bleed events observed persist or integrate into lasting behavioral change. The proof-of-concept argument is that the mechanisms are present and that the encounter is productive---not that the format is finished or that the theoretical extensions claimed here are definitive. They are the product of a first-contact experiment, offered as precisely that.

The questions the thesis opens are as follows. If alibi depth determines whether a role channels or buffers bleed, are there principled criteria for choosing which function a given role should serve in a given format? Is there a systematic relationship between transformation goal and required alibi type? If procedural rhetoric operates through incentive logic rather than structural symbolism, how should designers audit their formats for divergence between encoded and incentivized worldviews---and what does it mean to ``align'' them when transformative and entertaining logics may pull in different directions? If the transformational container has depth, what does a well-specified utilization level look like, and are there generalizable design patterns for growth legibility across different transformative game formats? And if the prior status problem is structural rather than psychological, does transformative game design applied to existing communities require a staging principle---attending first to participants for whom transformation is most available, and using the resulting community norm shifts to create conditions for broader change?

These questions are not limitations of the thesis. They are the generative output of applying a framework rigorously to a context it was not built for. The intersection of game design scholarship and forensics pedagogy is productive not because debate provides a convenient application site for theories developed elsewhere, but because the encounter reveals what those theories assumed without knowing it---and equips the field to design more precisely as a result.
