\section{INTRODUCTION}

\subsection{Brief Overview}

Through this thesis project, I apply transformative game design theory and best practices to create a new competitive debating format. This research employs Research through Design (RtD) methodology to develop and playtest an alternative debate format that addresses what I believe to be systematic problems in widely-used formats, particularly World Schools and British Parliamentary debate.

Structurally, I define how the major debate rulesets (called formats) are already games, are transformative and are roleplaying. Then I argue that there are issues with them that prevent them from being as positively transformative as they should be, and use transformative role-playing game design theory to point how they ought to change to address said issues. In the methods section I explain how I go about designing and testing a new debate format that should address these issues. Based on the same theory I defend my design and describe how  it evolved in light of playtesting. 
 
\subsection{Motivation}

I approach this research as a complete community member with 15 years of experience in competitive debate across multiple roles: debater, judge, coach, and organizer. This insider perspective informs my reformist motivation to improve formats that have both shaped and limited the community I value. I believe debate is a class of role-playing games with immense transformative potential, yet the design of major formats remains calcified, with game design theories and practices entirely absent from attempts to improve them.

\subsection{Problem Statement}

Although a very diverse range of debate formats are played all over the world, the practice seen in other games of iterative improvement of the game based on discoveries in game design is entirely absent, most formats being deeply calcified. Their competitive communities evolve new "metas", modes of play and inpterpretations of rules, without game design supervision. These are poised to lead, in time, to a pattern of accepting instead of solving the issues in the design of these formats. For example, based on my fifteen years of observation, participation, and informal conversations with fellow practitioners, I believe and hypothesize that the two most popular debate formats, World Schools and British Parliamentary, whose communities in Europe are deeply intertwined, exhibit the following problematic tendencies:

First, they seem to push debaters towards value convergence, potentially causing them to become less tolerant of people with different values and diminishing their ability to convince people from different backgrounds. Second, they appear to teach debaters to speak very fast and with extreme information density, which could diminish their ability to meaningfully support their positions to people not trained as debate judges. Third, in my opinion they push communities towards competitive hierarchical behavior that undermines psychological safety.

These observations, while grounded in extensive practitioner experience, have not been systematically validated through empirical research and this thesis does not intend to extensively prove them, although a later section will present their echoes in literature. I will try throughout this thesis to design a format that addresses these hypothesized problems, and through my design process, to generate evidence about how these problems are triggered by and are solvable through the design of the game.

\subsection{Research Question}

How can the application of transformative game design theory inform the creation of a competitive debate format that manages to elevate its transformative potential (improved critical thinking, improved discourse, self empowerment etc) by addressing the issues of value convergence, discourse speed and community hierarchy?

\subsection{Scope and Limitations}

This thesis aims to demonstrate proof of concept: that alternative debate formats can be systematically designed to address problems, and that such designs produce observable and articulable differences in participant experience. In particular it will try to demonstrate how a debate format can address one or more of the identified problems without sacrificing entertainment value, competitive appeal or educational potential.

The thesis does not aim to produce robust claims about long-term educational efficacy or entertainment value, as such claims would require longitudinal studies with larger samples than a master's thesis permits. Its main contribution to knowledge is the simple proof that debate formats can be designed better if looked at as roleplaying games. That is why the research involves approximately 4 playtest sessions over two months with experienced debaters from the Romanian debate community. One such session has already been held and a second design iteration drafted; the remaining sessions will culminate in qualitative analysis of observational and interview data sufficient to demonstrate how the design methodology employed can be systematically applied to improve the debate landscape.
