\section{METHODS}

\subsection{Design Philosophy}

This research employs Research through Design (RtD) as primary methodological framework. RtD positions the design process itself as inquiry, where knowledge emerges through making and iterating upon designed artifacts \parencite{coulton2017games,zimmerman2007research}. The design process follows \textcite{zimmerman2003play} iterative methodology, treating play as research. Each playtest generates data informing subsequent decisions, creating a spiral of refined understanding.

\subsection{Development Process}

Development unfolds through three cyclical phases. In the generation phase, I explore potential mechanics addressing theoretically-identified problems, using brainstorming sessions, design conversations with debate peers, and analysis of existing formats. In the critical scrutiny phase, I evaluate whether candidate designs congruently address problems without introducing new difficulties (e.g., judge cognitive load, time limitations), using the theoretical framework as evaluation criteria. In the validation phase, surviving elements undergo increasingly rigorous testing: informal design conversations with experienced debaters, small pilot playtests identifying mechanical problems, and a larger playtest as a mini-competition with structured observation and interview protocols.

These phases spiral until the format demonstrates sufficient coherence, playability, and alignment with transformational goals, or allocated resources run out. The first playtest (one debate) was held in March 2026 at the Jean Monet debate club in Bucharest, with senior debaters aged 18--20. A second design iteration was drafted based on its findings. The second playtest was conducted in April 2026 with seven veteran debaters in a familiar informal setting, yielding a third design iteration. Both sessions are documented in Annexes A--G.

\subsection{Research Methods}

For analyzing the quality of the design, two playtests were organized. During these sessions the research methods used were participant observation and structured post-session debriefs, followed by qualitative data analysis.

\subsubsection{Participant Observation}

Following \textcite{baker2006observation}, I adopted a participant-as-observer role during both playtest sessions. My researcher status was known to participants, and I participated actively in the second session to obtain firsthand understanding. When observing debates where not participating, I functioned as observer-as-participant: present and known but not actively engaged.

Observation focused on overall debate flow and more granularly on behaviors relevant to entertainment value, rule legibility and adherence ease, and elements distinct from standard formats, judging whether these indicate positive or negative tendencies vis-\`{a}-vis identified problems and the theoretical framework's predictions.

Informed by the advice of \textcite{spradley1980participant}, I recorded field notes in two phases: condensed and expanded. Condensed notes, taken during debates, captured key moments, quotations, and behavioral observations using structured protocols. Expanded notes, written within 24 hours of the sessions, elaborated, added contextual observations, and recorded analytical reflections. Both sessions were audio-recorded with participant consent to enable field note verification and capture verbal content not fully documented in real-time.

Observation spanned two structured playtest sessions over two months.

\subsubsection{Participants}

Debaters were recruited for both sessions with the aid of ARDOR Muntenia, the NGO managing the debate community in Bucharest and the south of Romania. They directly inquired various clubs about including a playtest as part of their weekly activity and issued open calls for debaters in organizing the sessions.

In Romania, the young adults (not minors) debate community is quite homogenous regarding debate experience, as almost all are senior debaters with training experience. Many of these are students or professionals living in Bucharest. The sample answering my open call will likely be representative in debate experience for the most probable target outside this research: experienced, not novice, debaters.

The majority of Romanian debaters are minors. Ideally, the study would involve minor participants, but ethical guidelines pushed toward this strategy as best alternative.

The two sessions involved a total of approximately fifteen participants. This sample size is consistent with the proof-of-concept nature of the research: the goal is demonstrating that design-based interventions can produce articulable differences in experience, not establishing generalizable population-level effects.

\subsubsection{Post-Session Debriefs}

Structured post-session debriefs complemented observation by accessing perspectives not directly observable. Following each session, participants completed a structured debrief form (Annex C) covering general experience (positive/negative, contrast with other formats, perceived difficulty), intentions to play in this format again, and reflections on specific mechanics. The debrief form was developed based on the theoretical framework's predictions about transformative mechanisms.

Debrief discussions occurred immediately following each session, ensuring fresh recall. They were conducted in person and supplemented by the researcher's reflexive notes recorded within 24 hours.

\subsection{Data Analysis}

Analysis follows a design-oriented comparative approach. The primary purpose is not generating generalizable theory but informing ongoing design decisions and articulating how the designed format produces different experiences than existing formats. This aligns with experimental game design methodology \parencite{waern2015experimental}, where design experimentation becomes research when ``the aim is to understand something generic, some more fundamental aspect of game design.''

Before data collection, the five evaluation fields derived from the research question and theoretical framework were operationalized into a taxonomy of \textit{potential clues}: specific observable behaviors or reported experiences that would count as evidence for or against the format's intended effects in each field. Both positive clues (behaviors consistent with the format achieving its intended effect) and negative clues (behaviors indicating a design gap or failure to achieve it) were specified at this stage. This pre-specification ensures the analysis remains responsive to disconfirming evidence rather than converging only on confirming observations. The complete taxonomy, reproduced in Annex~I, thus defines in advance what would need to be observed to consider the format effective or ineffective in each field.

Between playtest sessions, rapid analysis focused on actionable insights: what mechanics worked and should be preserved, what did not work and caused confusion or disengagement, and what surprised me and suggested new opportunities. This treats playtesting as ``play as research'' \parencite{zimmerman2003play}, with each session generating questions informing the next design cycle.

After the second playtest, systematic comparative analysis proceeded through three phases. First, initial coding following \textcite{lankoski2015formal} identified and labeled data segments describing participant experiences, behaviors, and reflections with descriptive rather than interpretive codes. Second, comparative code analysis organized codes around continuity (experiences similar to traditional formats), difference (experiences diverging from them), and novelty (experiences with no parallel in standard formats). This comparative structure also informs evidential weighting: novel and different behaviors constitute stronger design evidence than continuities, which may reflect participants' carry-over training rather than effects of the new format. Third, coded segments were mapped onto the potential clue taxonomy, confirming or failing to confirm each clue; confirmed clues were then connected to the specific format mechanics that appear to produce them---the causal link recorded as a design principle in each Annex~I entry.

This approach can demonstrate: (a) the designed format produces experiences participants describe as different; (b) specific mechanics appear connected to specific differences; and (c) differences align with intended goals. It cannot demonstrate persistence, superiority, or generalizability, appropriate limitations for proof-of-concept Research through Design where contribution lies in demonstrating that systematic design can address identified problems and in articulating design principles meriting further investigation or usage.

\subsection{Ethical Considerations}

This research adheres to Uppsala University's ethical guidelines for research involving human participants. All playtest participants were informed of the purpose and context, how data would be gathered, stored, and analyzed, both during and after recruitment. All were able to offer and retract consent, granularly or wholesale, to participation in every process part and to data being collected, stored, or used. Peers whose input was used in developing this format were informed of context and credited. All data is stored securely in encrypted Microsoft OneDrive offered by Uppsala University, separated by content type, and will be deleted once thesis development and presentation is complete. All data has been anonymized and is not publicly available.

To ensure psychological safety, the rigorous ethical standards imposed by the ARDOR Ethical Guide and ARDOR Trainer Safety Guide will be followed. Debate topics will avoid personal trauma topics and debaters will be able to refuse to engage with motions. Debaters will be able to always talk to a third-party equity officer regarding emotional safety issues and report abusive or distressing behavior. During debates, all debaters will be able to exit without explanation.

Beyond these, all participants will have pre-session briefing and post-session debrief providing chances to address potential distressing elements before activity and discuss experience afterwards.

\subsection{Positionality and Reflexivity}

I approach this research as a complete community member \parencite{adler1987membership}. My fifteen years in competitive debate have shaped not only my knowledge but my values, assumptions, and emotional investments. This informs several positions I must acknowledge: insider knowledge enabling nuanced understanding but potentially creating blind spots; reformist motivation emerging from genuine frustration with existing formats, predisposing me toward confirming my critique; a novel disciplinary lens through game design rather than the educational or rhetorical frameworks common in forensics scholarship; and power dynamics created by my status as former Team Romania coach and successful debater, which may influence participant responses. My competitive success within the formats I critique creates additional tension regarding whether my familiarity enables recognition of subtle issues or obscures problems I cannot fully perceive.

To support reflexive awareness, I maintained a reflexive journal and followed each debrief session with a reflexive note \parencite{knott2022interviews}, actively seeking data contradicting expectations. Given that some participants were known to me, recruitment was externalized to ARDOR Muntenia staff, and data from participants with close prior relationships was flagged while triangulating self-report data with observational data where possible.

\paragraph{Scope Note}
The original design of this study envisioned approximately four iterative playtest sessions. In practice, two sessions were completed within the thesis timeline. While this reduces the breadth of iterative feedback, both sessions yielded substantive observational and debrief data sufficient for proof-of-concept assessment. The implications of this reduced scope are discussed in the Results chapter.
