\section{RESULTS}
\label{sec:results}

The question admits two decomposed sub-questions. First: can a research method of applying transformative game design theory to this problem domain be demonstrated? Second: does that research method yield expected results? Each is addressed in turn.

\subsection{Demonstrating the Method: The Design Process as Contribution}

The first sub-question is answered by the thesis itself. The preceding chapters document a systematic progression from theoretical analysis (Sections~\ref{sec:background} and~\ref{sec:theory}), through iterative format design grounded in design theory principles (Section~\ref{sec:design}), to empirical testing and reflection. This sequence constitutes a worked example of how transformative game design theory can be operationalized in a competitive debate context. The design decisions recorded in Annexes A, B, and G---and the reasoning connecting each iteration to observed player behavior---demonstrate that such an application is coherent, achievable, and produces testable outcomes. Whether or not the resulting format succeeds transformatively, the method of application has been rendered explicit and repeatable.

\subsection{Evaluating the Results: Evidence from Playtests}

The second sub-question concerns whether the designed format shows signs of achieving its intended effects. Two playtests were conducted and analyzed using the clue-taxonomy method described in Section~\ref{sec:methods}; the complete taxonomy is in Annex~I. The following analysis discusses the most significant confirmed clues across the five evaluation fields. Behavioral and emergent evidence is weighted above self-report throughout, and social desirability effects are treated as a particular concern for Field~1 (see Annex~I preamble).

\subsubsection{Field 1: Engagement, Appeal, and Long-Term Viability}

For a debate format to sustain a community, it must be sufficiently engaging to generate voluntary, repeated participation. Several clues point in this direction, though the social desirability caveat above applies with particular force here.

Self-reported engagement was consistently positive across both playtests: participants described the format as ``surprisingly liberating,'' reported ``a lot less anxiety,'' and called the Huddle phase ``very very fun.'' These reports constitute moderate evidence only, given the playtest context. The stronger engagement signal is behavioral: in the second playtest, participants spontaneously adopted the term ``inner voices''---a term absent from the ruleset---to refer to each other during play. Organic language generation of this kind suggests genuine imaginative investment rather than performed enthusiasm, and is coded as a strong clue in Annex~I.

A recurring negative engagement clue was the scripted Rolling Out ritual, reported as ``very robotic'' in the first playtest and producing mumbling and requests for justification in the second. This is best read not as a wholesale failure of the format's immersive premise, but as evidence of the genre alibi that debate as a practice offers: ritual language that would be natural in a TTRPG sits uneasily in a debate context, because debate's social contract frames excessive ceremony as frivolous. Design iteration has continued to address this---the ritual was revised in the third iteration (Annex H), though that version remains untested.

The differential appeal across participant profiles is a relevant secondary finding. Women and debaters who had been less successful in traditional formats expressed consistent preference for Inside Out across both sessions; the one participant who found the format unsatisfying was a highly competitive debater who missed the ``ego-boosting'' experience of decisive victory. Whether this profile-specific response represents a limitation or a design confirmation depends on the format's intended audience.

Section~\ref{sec:design} noted format novelty as a potential barrier to entry. The 40-minute brief required in both sessions---even with experienced debaters---confirms that the learning curve is real. However, both sessions produced substantively engaged play within a single session, suggesting the cost is one of onboarding rather than fundamental inaccessibility. Each design iteration has addressed specific gaps that generated confusion in the previous session.

\subsubsection{Field 2: Productive Pressure Mechanisms}

Competitive debate's transformative potential depends partly on pressure: the cognitive and social demands of the activity push participants to grow. A format without sufficient challenge risks becoming a low-stakes social exercise without developmental benefit. Clues from both playtests suggest the format redirects rather than removes intellectual pressure, though this redirection also surfaced a significant design gap.

Participants in both playtests engaged with high effort: in the first session, players constructed elaborate and nuanced solutions, made careful value-priority arguments, and delivered some notably passionate speeches---try-hard investment by any standard; in the second, players spontaneously produced contextually rich, roleplay-oriented speeches with worldbuilding detail, none of which was instructed. Participants in the first playtest described ``architecting solutions'' as ``very complex and interesting strategically,'' a self-report consistent with the behavioral evidence. Both sessions also produced strikingly high levels of acting and role-play, raising the question of whether emotional expressiveness came at the cost of analytical rigor; the evidence suggests it did not, as elaborate solutions, practicalities-based reasoning, and careful value hierarchies remained consistently present alongside the emotional engagement.

A notable finding about the source of performance pressure: across both sessions it became clear that the mechanic driving effort was not the win/loss dynamic (which is absent) nor primarily the commendation system, but a social one---participants wanted to be recognized and appreciated by their fellow Inner Voices and by the Subject for the quality and resonance of their contributions. This directly addresses the design tension raised in Section~\ref{sec:design} about whether competitive motivation can survive without win-loss outcomes: the answer appears to be yes, but via a different mechanism than the commendation architecture alone---social recognition within the group shifts competitive stakes from zero-sum outcome-seeking toward collaborative reputation, a healthier motivator that is also more aligned with the format's transformative intent.

A notable negative clue was Subject role confusion: in the first playtest the Subject player described not knowing ``what to say,'' uncertain how to hold emotional investment and evaluative neutrality simultaneously, producing visible hesitation. This was a specific design gap in the S0 and S1 instructions, addressed in the second design iteration; the confusion did not recur in the second playtest. The second session also offered a partial answer to the design tension about whether the Subject's cognitive burden is productive challenge or overwhelming difficulty: the Subject who navigated the role most successfully was the one who leaned into roleplay rather than maintaining analytical detachment---simply inhabiting the character, responding emotionally, and asking questions from genuine curiosity resolved the cognitive tension of trying to reconcile the roleplay and evaluative functions simultaneously. This is consistent with the broader argument in Section~\ref{sec:background} that debate is fundamentally a form of role-playing and cannot be done well without inhabiting the role.

The commendation system remains a partially open question. Evaluators in both playtests found commendations difficult to assign consistently, pointing to a pressure-and-recognition mechanism that is conceptually coherent but calibrationally immature at small informal scale. However, the Evaluator in the second session noted that at larger circuit scale---where the commendation list is memorized over repeated use and Evaluators can reserve higher-tier commendations for truly outstanding moments---the system would feel considerably more organic; at small scale, a practical workaround is to observe what debaters do well and work backwards to match a commendation retrospectively rather than scanning the list prospectively.

\subsubsection{Field 3: Growth Direction---Useful Skills and Perspectives}

A central design goal was to redirect the skills debate develops toward more transferable outcomes: persuasive communication, perspective-taking, contextual reasoning, and collaborative discourse. The clues here are mixed: strong behavioral evidence in the second playtest, moderate self-report in the first, and one significant negative clue in the first.

In the first playtest, participants articulated learning to ``think about the context of decisions,'' to be ``more human and more honest about human nature,'' and to recognize that ``sometimes decisions are personal, not universal.'' These self-reported reflections are moderate evidence: they align with the format's procedural rhetoric \parencite{bogost2007persuasive} and are on-target with design intent, but they are retrospective and unverified by pre-session baseline. A behaviorally notable pattern across both sessions is that play diverged increasingly from standard debate conventions as each session progressed: early in the debate participants applied familiar genre habits, but speech by speech they adjusted toward the format's distinct demands---a natural discovery arc that suggests the format's different expectations become legible through play and that skill development in the new directions will follow with repeated exposure. Equally important is what was absent: no non-standard or unhelpful new play behaviors were rewarded, and no performative hollow emotionality was observed---negative clues that, had they manifested, would have undermined the growth-direction claim (see Annex~I).

The second playtest provides the strongest positive clues for this field. All players immediately adopted a roleplay-oriented stance---second-person address, direct eye contact with the Subject, contextual worldbuilding detail---none of which was instructed in the ruleset. This spontaneous behavior is behavioral rather than self-reported, and coded as a strong clue in Annex~I: it suggests the format's structure elicits perspective-taking without instruction, though this observation is confined to a single session and cannot yet be called reliable.

A further behavioral cluster that emerged in the second playtest is worth noting for this field: players spontaneously adopted a polite and collaborative form of counter-argumentation---shifting focus and building on others' contributions rather than dismantling them---and the second round of speeches showed organic movement toward a common solution, with players converging rather than intensifying conflict. These patterns are novel relative to standard debate and are among the clearest instances of the format's procedural rhetoric encoding collaborative negotiation as a transferable skill.

\subsubsection{Field 4: Community Dynamics and Reduced Toxicity}

One of the design's primary goals was to reduce the hierarchical and adversarial dynamics that contribute to attrition and harm in competitive debate communities. This field presents both the most consistent positive clues and the most notable pattern of absent negative clues.

Reduced anxiety was reported consistently and independently across both playtest groups---``much much less anxiety-inducing'' than standard competitive debate, a ``feeling of salvation regardless of result,'' the sense that ``everyone distinguished himself by his own style.'' The multi-session consistency strengthens confidence in this finding beyond what a single playtest would allow. One honest caveat is offered: the format's reduced established status means less social capital is staked on performance, so some portion of the anxiety reduction may not persist in a mature Inside Out circuit. This is an empirical question for future study.

A strong behavioral clue from the second playtest is the emergent norm that ``outright attacking others' points feels wrong, disrespectful, and unnecessary.'' This was not stated in the ruleset; it arose from the format's structural logic---a procedural rhetoric outcome in the precise sense theorized by \textcite{bogost2007persuasive}. That players found collaborative framing more natural and more effective than adversarial rebuttal suggests a self-reinforcing alignment between competitive incentives and transformative design intent.

Both sessions produced instances of brave, unconventional play that was acknowledged and positively received. In the first playtest, the C team adopted a deliberately intimate and conversational delivery---emotionally vulnerable and unhurried---that is highly uncommon in the standard formats those participants were trained in, and it was recognized in the oral critique. In the second playtest, brave play took multiple forms: one player opened their subject speech with a background soundtrack playing from their phone; the C team built their entire case around a worldbuilt character (the Subject's mother) and her emotional stakes; players throughout the room adopted expressively roleplay-heavy delivery, and the room responded with visible gesture-based approval---laughing, nodding, leaning forward---throughout. This atmosphere of genuine appreciation for creative risk-taking is a strong behavioral clue for the evaluative culture the design intends to cultivate.

Equally informative is what was not observed. None of the three negative clues for this field---expressions of distress, arrogant or unempathetic behavior, or praise of such behavior---manifested across either playtest or in the transcripts. Given the emotional intensity of some moments in the second session (including an unplanned bleed event where a participant shared privately that the scenario resonated with a real-life situation they were facing that same day), the absence of distress is a meaningful signal that the debrief structure provided sufficient containment.

\subsubsection{Field 5: Epistemic Humility}

The deepest transformative aim of the format is to cultivate epistemic humility: a disposition to take multiple perspectives seriously, to recognize that different value frameworks can each be valid, and to hold one's conclusions with appropriate uncertainty. This field contains some of the strongest behavioral clues in the dataset, though all are from a single playtest session.

In the first playtest, one participant explicitly articulated: ``every team says valuable stuff and it is important more to see what is important to us.'' This moderate clue is notable for being unprompted, and for capturing precisely the attitudinal shift the format targets. Further supporting evidence from the first session includes participants recognizing that arguing someone is ``morally wrong'' for a personal decision feels ``a bit absurd'' in this format, and the observation that the format makes personal value priorities legible rather than evaluating them against an external standard.

The strongest clues come from the second playtest. The spontaneous drift toward synthesis in the second round of Inner Voice speeches---players converging rather than intensifying conflict, despite no rule incentivizing this---is a behavioral, transcript-confirmed, emergent indicator that the format's procedural logic orients participants toward integration rather than dominance. A notable related finding is that players in the second session organically renamed themselves ``Inner Voices'' during play---a label that positions them as parts of a shared internal deliberation rather than opposing sides, colleagues rather than competitors. This renaming was subsequently formalized in the third design iteration.

The second playtest debrief also generated a theoretically rich conversation about how the S0 speech institutionalises the practice of surfacing personal biases openly. Participants contrasted this with standard formats, in which the pretense of objectivity is maintained throughout: those formats, they argued, falsely presuppose that bias can be excluded, propagating the notion that only one type of ``objective'' person can judge well and legitimizing social pressure to perform that identity. The S0 mechanism, by contrast, frames bias as perspective rather than contamination---a structural move toward epistemic humility that is enforced by the format before any argument is made.

The unplanned bleed event---a participant sharing privately that the scenario resonated with a real-life situation they were facing that same day---is the single highest-strength clue across the entire dataset: it was spontaneous, private (not performed for the researcher), and constitutes the transformational mechanism functioning exactly as theorized in Section~\ref{sec:theory}.

No negative clues for this field---unequivocal condemnation of positions or belief-holders, or praise of such condemnation---were observed across either playtest or the transcripts. The absence is meaningful given that both motions used involved choices with genuine ethical weight.

\subsection{Proof of Concept Assessment}

Taken together, the clues documented across both playtests provide reasonable grounds for a proof-of-concept claim. The format:

\begin{itemize}
    \item was experienced as engaging and appealing, particularly by participants underserved by traditional formats;
    \item maintained sufficient intellectual pressure through a redirection rather than removal of challenge;
    \item elicited spontaneous perspective-taking, value exploration, and contextual reasoning;
    \item generated emergent community norms oriented toward collaborative rather than adversarial discourse; and
    \item produced instances of epistemic humility, bleed effects, and personal value resonance consistent with the theoretical framework's predictions.
\end{itemize}

No single piece of evidence is conclusive. Small sample sizes, the absence of longitudinal follow-up, and the researcher's dual role as designer and observer all limit what can be inferred. However, the convergence of multiple independent clues across two separate playtest groups---each producing similar patterns of engagement, anxiety reduction, and spontaneous norm formation---provides a coherent picture that justifies continued investigation. The format warrants further empirical study with larger groups, more experienced facilitators, and longitudinal measurement of the skill and attitudinal outcomes theorized here.

\subsection{Limitations}

The original design of this study envisioned approximately four iterative playtest sessions. Two were completed within the thesis timeline. While this reduces the number of design iterations and the breadth of participant exposure, both sessions yielded substantive observational and debrief data sufficient for proof-of-concept assessment. Future work should extend the iterative cycle.

The researcher's insider position as a fifteen-year competitive debater and the format's designer is both a strength (enabling nuanced interpretation of participant behavior) and a limitation (creating potential for confirmation bias in clue selection and interpretation). Reflexivity notes were maintained throughout, and the Annex~I clue list explicitly notes the evidential strength and directionality of each inference.

Finally, both playtest groups consisted of experienced debaters already familiar with competitive formats. The format's performance with novice participants---for whom the contrast with standard debate would be less salient---remains untested.
